---
title: "Single-cell RNA-seq"
subtitle: "From counts to clusters · BINF90004 Bioinformatics Case Studies"
author: "Dr Jiadong Mao · Melbourne Integrative Genomics"
date: today
format:
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true          # press 'b' to draw, 'c' for chalkboard
    incremental: false
    width: 1280
    height: 720
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
suppressPackageStartupMessages({
    library(Seurat); library(ggplot2); library(dplyr); library(patchwork)
})
# Resolve the data folder for every layout this deck is rendered from:
# repo root, repo slides/, and the private authoring folder (../dist).
D <- if (dir.exists("data")) "data" else
     if (dir.exists("../data")) "../data" else "../dist"
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")

# Loaded here rather than mid-deck: two later slides use it, and if it is
# loaded inside a slide chunk, deleting that slide silently breaks them.
seu <- readRDS(file.path(D, "P1_start.rds"))
```

# Part 1 {.center}

## What does a bioinformatician do now? {.center}

::: {.notes}
12 minutes. This is not a technology showcase - it is the reason the practical
is built the way it is. Do not rush it.
:::

## A show of hands

**Who has used an LLM to write analysis code in the last month?**

<br>

**Who checked what it chose for you?**

::: {.notes}
Expect nearly every hand for the first, very few for the second. That gap is
the whole lecture.
:::

## This is a product category now

::: {.incremental}
- **Claude Science** (beta, 2026) — multi-agent workbench for scientific analysis
- Connectors: Benchling, 10x Genomics, PubMed, BioRender, Synapse
- **Agent Skills** — packaged, reproducible analysis workflows
:::

<br>

::: {.fragment}
The first Agent Skill Anthropic released is **`single-cell-rna-qc`**.

It automates the quality control step of scRNA-seq.

**That is section 2 of your practical this afternoon.**
:::

## It is genuinely good

**BioMysteryBench** — expert-written questions on real datasets, objective
ground truth. You are given the data files and a question; looking the source
dataset up is against the rules.

| | Accuracy |
|---|---|
| Human-solvable tasks | **~77%** |
| Human-*difficult* tasks | **~23.5%** |

Models solved some problems expert panels could not.

<div style="font-size: 0.5em; color: #5f6b7a; margin-top: 0.8em;">
Anthropic (2026), *Evaluating Claude's bioinformatics research capabilities
with BioMysteryBench*. Problems: <https://huggingface.co/datasets/Anthropic/BioMysteryBench-preview> (CC-BY-4.0)
</div>

::: {.notes}
Accuracy figures are Anthropic's April 2026 release, measured on the original
99-problem set (76 solvable / 23 hard). PLAN.md attributes them to Opus 4.6;
some secondary coverage attributes the same two numbers to Sonnet 4.6 - the
source page presents them as a graph, so CHECK THIS before delivery.

The benchmark was revised in July 2026 (v11): 99 -> 90 problems, 73 solvable
/ 17 hard. Nine were withdrawn after a June audit found the answer keys wrong
or the question underivable from the data provided. That is worth a sentence
out loud - the benchmark's own ground truth needed checking too.

See PLAN.md 3.5.
:::

## What the questions actually look like

::: {.columns}
::: {.column width="50%"}
**Mostly gets right** — solves ~77%

<div style="font-size: 0.72em; line-height: 1.25;">
*"What organism does this crystal structure belong to?"* → `Homo sapiens`

*"What bacteria is found in this sequenced dataset?"* → `Bacillus licheniformis`

*"Identify the transcription factor whose binding sites these peaks
represent."* → `CTCF`
</div>
:::
::: {.column width="50%"}
**Mostly gets wrong** — solves ~23.5%

<div style="font-size: 0.72em; line-height: 1.25;">
*"How many tissue groups are collected, and from how many different
species?"* (metagenomes, non-human hosts) → 2 tissues, 3 snake species

*"What is the most common microRNA associated with the genes given in this
data?"* → any `hsa-let-7` family member
</div>
:::
:::

::: {.fragment}
On the hard ones it fails **about three in four**.
:::

<div style="font-size: 0.45em; color: #5f6b7a;">
Questions verbatim from the public 5-problem preview, v11 (July 2026); answers
are the published rubrics. Which individual questions Claude failed is not
published — only the per-split rates.
</div>

::: {.notes}
The point of this slide is the shape of the questions, not the answers. Every
one is "here are the files, tell me what they are" - no method is specified,
so the model has to choose the whole pipeline. That is exactly the position
you are in this afternoon.

Note what separates the two columns. The left ones have a single determinate
answer reachable by one canonical route: BLAST it, motif-scan it. The right
ones need several correct choices chained together, and each has a count or a
family in the answer - more places to be almost right. This is the same gap
you will see in the pre-run: it does the standard move well and the judgement
calls invisibly.

Be precise if asked: the right-hand column is the *class* it mostly fails,
not two questions it is known to have failed. Anthropic publishes the split
rates, not per-question results, so nobody can say which individual ones it
got wrong. Worth saying out loud - it is the same "check what is actually
being claimed" move the whole lecture is about.

Do not spend the 1-2-of-5 brittleness finding here - it is the whole of
"Would it do the same thing twice?" at the end of Part 5.

If asked why snakes: the data is anonymised, so the model has to identify
three host species from metagenomic reads without looking up the accession.
:::

## So I gave it our dataset

I uploaded the folder you are about to analyse. **No question, no instructions**
— just the files.

<br>

100 minutes later, unattended: QC, integration, clustering, annotation,
composition, differential expression, four figures, four tables, a processed
data object and a written report.

::: {.fragment}
Cost: **$7.09**.
:::

::: {.notes}
Full scoring in prerun/SCORING_run_A0.md. Every quote from it in this deck is
verbatim, but they are now spread across the morning rather than all shown
here - see the teaser slide's note for where they land.
:::

## It got a lot right, and two things badly wrong

<br>

I will not tell you which yet — you cannot judge the calls it made until you
know what the calls *are*.

::: {.fragment}
<br>

So we will do it the other way round. Each time you learn a step this morning,
we will look at what the agent did with that step.

**By the end of today you will be able to score it yourself.**
:::

::: {.notes}
This replaces the old front-loaded critique block. Do not name the two failures
here - composition testing and per-cell DE are both L2 material, and naming
them now spends the payoff L2 builds to in "What the agent did, scored".

The promise made here is kept four times: after Cells or nuclei?, after QC
metrics, after Integration, and after Resolution.
:::

## But what about taste?

**Chen, Zhao & Cohan (2026)** — *Measuring the Gap Between Human and LLM
Research Ideas*, arXiv:2607.01233

They do not score ideas one at a time. They reconstruct the literature context
for **11,683 papers** and compare the *distribution* of human vs LLM ideas.

::: {.fragment}
**Any single LLM idea looks fine. The narrowness only shows up in aggregate.**
:::

## The gap

| | Human | LLM |
|---|---|---|
| "Bridge-like" opportunities (connect A to B) | **12.1%** | **47–64%** |
| Synthesis / unification methods | **5.1%** | **22–39%** |
| Diversity (normalised entropy) | **0.926** | **0.737** |

::: {.fragment}
And extended reasoning made it **worse** — one model went from 49.7% → **71.1%**
bridge-like with thinking mode on.
:::

## Why you should care

LLM research ideas cluster on:

- *"combine X with Y"*
- *"apply method A to dataset B"*

<br>

::: {.fragment}
*"Apply a foundation model to spatial transcriptomics."*

*"Integrate multi-omics with a graph neural network."*

<br>

Recognise those? Some of you have written them in a project proposal.
:::

::: {.notes}
This is the moment the room goes quiet. Let it.
Be fair to the paper: STEM-only corpus, one-shot prompting, reconstructed
contexts lack tacit knowledge. The claim is "current LLMs, prompted this way,
produce a narrow distribution" - NOT "AI cannot have taste."
:::

# Part 2 {.center}

## Why single cell {.center}

## Why RNA

DNA → **RNA** → protein

- DNA is the same in (nearly) every cell
- Protein is hard to measure at scale
- RNA is what the cell is *doing right now*, and we can count it

## Bulk vs single cell

**Bulk RNA-seq**: average over millions of cells

::: {.fragment}
A tissue that is 70% cardiomyocytes and 5% immune cells gives you a
cardiomyocyte-flavoured average, and the immune biology is invisible.
:::

::: {.fragment}
**scRNA-seq**: one measurement per cell. You can ask *which* cells, not just
*how much*.
:::

## How it works

1. Dissociate the tissue into single cells
2. Capture each cell in a droplet with a barcoded bead
3. Every transcript from that cell gets the same barcode
4. Sequence everything; sort by barcode

::: {.fragment}
**Output: a genes × cells matrix of counts.**
:::

## Cells or nuclei?

::: {.columns}
::: {.column width="50%"}
**scRNA-seq**

- whole cells
- needs tissue that dissociates gently
- cytoplasmic RNA included
:::
::: {.column width="50%"}
**snRNA-seq**

- nuclei only
- works on frozen / tough tissue
- **cytoplasm discarded**
:::
:::

::: {.fragment}
Adult heart muscle does not dissociate nicely. **Today's data is snRNA-seq** —
which is why `percent.mt` behaved the way it did ten minutes ago.
:::

## The first thing it said {.center}

Back to the agent I handed this dataset to. **This was the opening of its
report** — before any of you had a word for it.

> "mitochondrial fractions are uniformly low (donor medians 0.10–1.78%) **and**
> so are ribosomal fractions (0.4–1.3%). That combination points to a
> **single-nucleus prep** rather than whole cells, which changes how you'd handle
> ambient RNA."

::: {.fragment}
Nobody told it this was snRNA-seq. **It worked that out from the data**, and set
its mitochondrial threshold accordingly.
:::

## The matrix

```{r matrix-shape}
cat("genes:", nrow(seu), "\n")
cat("nuclei:", ncol(seu), "\n")
m <- SeuratObject::LayerData(seu, assay = "RNA", layer = "counts")
cat("zeros:", round(100 * (1 - length(m@x) / (nrow(m) * ncol(m))), 1), "%\n")
```

<br>

High-dimensional, and **overwhelmingly zero**. Some of those zeros are
"not expressed"; some are "expressed but not captured". You cannot tell which.

# Part 3 {.center}

## The workflow {.center}

## Three stages

**Raw data processing** → align reads, produce the count matrix

**Pre-processing** → QC, normalise, select features, correct batch, reduce
dimensions

**Downstream analysis** → cluster, annotate, test hypotheses

<br>

Visualisation throughout. We start at stage 2 — the count matrix is given.

::: {.notes}
Luecken & Theis 2019 figure. Reuse from the 2025 deck.
:::

## Today's dataset

**Sim et al. (2021), *Circulation*** — human heart maturation

| Group | Donors | Age |
|---|---|---|
| foetal | f1–f3 | 19–20 weeks |
| young | y1–y3 | 4–14 years |
| adult | a1–a3 | 35–42 years |

::: {.fragment}
**Nine donors.** Remember that number — in the second lecture it turns out to be
the only number that matters.
:::

# Part 4 {.center}

## Pre-processing {.center}

## Quality control

Not every barcode is a good cell:

- **Empty droplets** — ambient RNA only; low counts, low gene number
- **Dying cells** — membrane compromised, cytoplasmic RNA leaks out,
  mitochondrial fraction rises
- **Doublets** — two cells, one barcode; look like a chimera

::: {.fragment}
Tools exist for each (CellBender, scDblFinder, DoubletFinder). None of them
removes the need to choose.
:::

## QC metrics

```{r qc-violin, fig.width=11, fig.height=4.5}
VlnPlot(seu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
        group.by = "sample", pt.size = 0, ncol = 3) &
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

**Look at the donors, not just the distribution.** They differ, and a single
global threshold lands differently on each.

## I had predicted it would get that wrong

Now you know what `percent.mt` is, here is the threshold the agent chose — and
the one I bet it would choose.

I expected `percent.mt < 5` — the whole-cell default from every tutorial.

::: {.fragment}
<br>

I had also written down the "expert" rebuttal:

> *Heart is mitochondria-rich. Cardiomyocytes are packed with mitochondria. An
> aggressive MT filter deletes the cell type the study is about.*
:::

::: {.fragment}
<br>

**That reasoning is also wrong.** Median `percent.mt` here is 0.25%, and
cardiomyocytes have the *lowest* of any cell type — because these are nuclei and
the cytoplasm has been washed away.

**Its reasoning beat mine.**
:::

::: {.notes}
Own this. Saying "I got this wrong and the machine didn't" is the most
persuasive thing you can do in this room, and it is true.
:::

## Normalisation

Cells differ in total counts for reasons that are not biological — capture
efficiency, sequencing depth, cell size.

::: {.incremental}
- **Log-normalisation**: divide by library size, scale, log. Simple, fast.
- **SCTransform**: regularised negative binomial; returns Pearson residuals.
  Handles the mean–variance relationship properly.
:::

::: {.fragment}
We use SCTransform today. Neither is "correct" — they answer slightly different
questions.
:::

## Feature selection

~19,000 genes. Most are uninformative:

- not expressed
- expressed everywhere at the same level (housekeeping)

<br>

Keep the **2,000–5,000 most variable** genes. Everything downstream — PCA,
clustering — runs on those.

::: {.fragment}
Note what this means: **a gene you filter out cannot be discovered.**
:::

## Dimension reduction: PCA

Still thousands of dimensions, and they are correlated.

**PCA** finds linear combinations capturing the most variance.

- Use the first 10–50 PCs for everything downstream
- Faster, and drops noise in the high components
- **Not** for visualisation — 2 PCs is not enough

::: {.fragment}
How many to keep? An elbow plot, and a judgement call. *(Decision 2.)*
:::

## Visualisation: UMAP and t-SNE

Non-linear projections to 2D. Good at preserving **local** structure — cells
that were neighbours stay neighbours.

```{r umap-demo, fig.width=9, fig.height=4}
s2 <- readRDS(file.path(D, "checkpoints/02_integrated_clustered.rds"))
DimPlot(s2, reduction = "umap", group.by = "group", cols = group_cols,
        shuffle = TRUE) + ggtitle("UMAP, coloured by developmental stage")
```

## UMAP: what it does not mean {background-color="#F5E9E9"}

::: {.incremental}
- **Distance between clusters is not meaningful**
- **Cluster area is not population size**
- Apparent gaps can be artefacts of the parameters
- Clusters can appear that are not there
:::

<br>

::: {.fragment}
It is a picture for looking at, not a measurement. Never test a hypothesis on
UMAP coordinates.
:::

## Batch effects

```{r batch, fig.width=10, fig.height=4.2}
p1 <- DimPlot(s2, reduction = "pca", group.by = "sample", shuffle = TRUE) +
    ggtitle("PCA — before integration") + theme(legend.position = "none")
p2 <- DimPlot(s2, reduction = "harmony", group.by = "sample", shuffle = TRUE) +
    ggtitle("After Harmony")
p1 + p2
```

Cells cluster by **donor** instead of by cell type. Each sample carries its own
technical signature.

## Integration

**MNN** · **CCA** · **Harmony** — all find a shared space where donors overlap.

::: {.fragment}
::: {.callout-warning appearance="minimal"}
**There is no measurement that tells you it worked.**

"The donors mix better" is also exactly what over-correction looks like — and
over-correction deletes real biology.
:::
:::

::: {.fragment}
Here the donors *differ by developmental stage*. Some separation between them is
the signal we came to study. We integrate by **donor**, not by **stage** —
deliberately.
:::

## It also did something our practical does not

You have just seen that integration can remove real biology along with the
batch effect. Here is how the agent checked whether it had:

> "Clustering on unintegrated PCA gave clusters that were stage-exclusive *and*
> donor-exclusive — 7 clusters were ≥90% a single donor. That can't distinguish
> development from batch. After Harmony: **0** donor-dominated clusters, while
> **10** clusters stay stage-restricted. Since those now draw from all donors
> within a stage, the restriction is developmental."

::: {.fragment}
It designed a test to tell correct integration from over-correction. Your
practical only warns you that the difference exists.
:::

# Part 5 {.center}

## Clustering {.center}

## Graph-based clustering

1. Build a **shared nearest neighbour** graph — cells connected by how many
   neighbours they share
2. Find **communities** in that graph

<br>

**Louvain** (Seurat default) · **Leiden** (faster, better-connected communities)

## Resolution

```{r res-counts}
res <- c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8)
n <- sapply(res, function(r) length(unique(s2@meta.data[[paste0("SCT_snn_res.", r)]])))
knitr::kable(data.frame(resolution = res, clusters = n))
```

::: {.fragment}
There is no correct value. It depends on the question — and today, on you.
*(Decision 3.)*
:::

## A warning you will verify this afternoon

The published analysis of this tissue describes **8 cell types**.

::: {.fragment}
At the *lowest* resolution above, you already get **12 clusters**.
:::

::: {.fragment}
<br>

**A cluster is not a cell type.** It is a community in a graph you built, using
genes you selected, in a space you chose the dimension of.
:::

## Would it do the same thing twice?

From Anthropic's own evaluation, on the hard problems:

> nearly half of the wins are paths the model *stumbles onto* rather than
> reproduces

::: {.fragment}
<br>

**I have run this once.** So I cannot tell you whether a second run picks the
same thresholds, the same resolution, or notices the nuclei at all.

Neither could you, from a paper that reports one run.
:::

::: {.notes}
HONEST STATUS: n = 1. Do NOT claim run-to-run variance until A1-A3 are done -
see prerun/SCORING_run_A0.md. If you do get the repeat runs before the session,
replace this slide with the actual comparison; it is much stronger.
:::

## So what is left for you?

::: {.incremental}
- Not knowing more statistics than it does. It diagnoses things we have not
  taught you yet, unprompted.
- Not knowing more biology. It identified the assay from the data, and I
  could not.
- **Reading to the end** — noticing which number it put in the headline, and
  which one it buried.
- **Noticing the test that was never run.** You will find one this afternoon.
:::

::: {.fragment}
<br>

Every one of those requires you to already understand the method well enough to
know what *should* have been there.
:::

## The deal for today

You will make **five decisions** this afternoon that no tool makes for you:

1. Where to set your quality thresholds
2. How many principal components to keep
3. How finely to cluster
4. What cell type a cluster is
5. How to test for differential expression

::: {.fragment}
Each time: choose, then **justify it to the person next to you**.

That is the assessable skill, and it is what your report is marked on.
:::

## What you will do now

::: {.callout-note appearance="minimal"}

## Practical 1 — 65 minutes
**Decision 1.** Where do you set your QC thresholds?

**Decision 2.** How many PCs?

**Decision 3.** How many clusters are real?
:::

<br>

There is one hard sync point. You are not expected to finish every chunk.

If your R is broken, you can still do all three — **come anyway.**

