---
title: "Single-cell RNA-seq"
subtitle: "Module I: From counts to clusters"
author: "Dr Jiadong Mao </br> Melbourne Integrative Genomics (MIG) </br> BINF90004 Bioinformatics Case Studies 2026"
format:
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true          # press 'b' to draw, 'c' for chalkboard
    incremental: false
    width: 1280
    height: 720
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
suppressPackageStartupMessages({
    library(Seurat); library(ggplot2); library(dplyr); library(patchwork)
    library(igraph)
})
# Resolve the data folder for every layout this deck is rendered from:
# repo root, repo slides/, and the private authoring folder (../dist).
D <- if (dir.exists("data")) "data" else
     if (dir.exists("../data")) "../data" else "../dist"
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")

# Loaded here rather than mid-deck: two later slides use it, and if it is
# loaded inside a slide chunk, deleting that slide silently breaks them.
seu <- readRDS(file.path(D, "P1_start.rds"))

## ---- Toy kNN graph, shared by four slides (UMAP, SNN, clustering, resolution)
# Same reason as above: built here so that deleting any one of those slides
# cannot break the others. Four groups, and group 4 sits close to group 2 --
# that near-pair is what merges at low resolution on the resolution slide.
set.seed(1); n_toy <- 15; K_toy <- 6
toy <- data.frame(
    x = c(rnorm(n_toy, 0, .55), rnorm(n_toy, 3.1, .55),
          rnorm(n_toy, 1.6, .55), rnorm(n_toy, 4.4, .55)),
    y = c(rnorm(n_toy, 0, .55), rnorm(n_toy, 0.3, .55),
          rnorm(n_toy, 2.7, .55), rnorm(n_toy, 0.9, .55)))
# k nearest neighbours of each cell, then a de-duplicated edge list
toy_nb <- t(apply(as.matrix(dist(toy)), 1, function(d) order(d)[2:(K_toy + 1)]))
ed <- unique(t(apply(do.call(rbind, lapply(seq_len(nrow(toy)),
                     function(i) cbind(i, toy_nb[i, ]))), 1, sort)))
ed <- data.frame(i = ed[, 1], j = ed[, 2])
# SNN weight: how many neighbours do the two endpoints have in common?
ed$shared <- mapply(function(a, b) length(intersect(toy_nb[a, ], toy_nb[b, ])),
                    ed$i, ed$j)
ed <- cbind(ed, x = toy$x[ed$i], y = toy$y[ed$i],
                xend = toy$x[ed$j], yend = toy$y[ed$j])

toy_g <- igraph::graph_from_data_frame(ed[ed$shared > 0, c("i", "j")], FALSE,
                                       data.frame(id = seq_len(nrow(toy))))
igraph::E(toy_g)$weight <- ed$shared[ed$shared > 0]
# Louvain is greedy and order-dependent, so seed it: the cluster counts are
# quoted in the slide titles and must not move between renders.
toy_cl <- function(r) { set.seed(42)
    factor(igraph::membership(igraph::cluster_louvain(toy_g, resolution = r))) }

toy_pal <- c("#4E79A7", "#F28E2B", "#59A14F", "#E15759", "#B07AA1", "#76B7B2",
             "#EDC948", "#FF9DA7", "#9C755F", "#BAB0AC")
toy_base <- function(p, ttl) p + coord_equal() + theme_void() + ggtitle(ttl) +
    theme(plot.title = element_text(hjust = .5, size = 14), legend.position = "none")
# w = TRUE draws edge width/alpha by SNN weight; w = FALSE draws plain kNN edges
toy_segs <- function(w = FALSE) if (w)
    geom_segment(data = ed, aes(x = x, y = y, xend = xend, yend = yend,
                 linewidth = shared, alpha = shared), colour = "grey60") else
    geom_segment(data = ed, aes(x = x, y = y, xend = xend, yend = yend),
                 colour = "grey70", linewidth = .3)
```

# Part 1 AI bioinformatician?

::: {.notes}
10 minutes. This is not a technology showcase - it is the reason the practical
is built the way it is. Do not rush it.

Part 1 is now about AI in science generally. The pre-run on OUR dataset is held
back until QC in Part 3, where the mitochondrial quotes land on a metric the
room has just seen. Do not trail it here.
:::

## A show of hands

**Who has used an LLM to write analysis code in the last month?**

<br>

**Who checked what it chose for you?**

::: {.notes}
Expect nearly every hand for the first, very few for the second. That gap is
the whole lecture.
:::

## This is a product category now

::: {.incremental}
- **Claude Science** (beta, 2026) — multi-agent workbench for scientific analysis
- Connectors: Benchling, 10x Genomics, PubMed, BioRender, Synapse
- **Agent Skills** — packaged, reproducible analysis workflows
:::

<br>

::: {.fragment}
The first Agent Skill Anthropic released is **`single-cell-rna-qc`**.

It automates the quality control step of scRNA-seq.

**That is section 2 of your practical this afternoon.**
:::

## It is genuinely good

**BioMysteryBench** — expert-written questions on real datasets, objective
ground truth. You are given the data files and a question; looking the source
dataset up is against the rules.

| | Accuracy |
|---|---|
| Human-solvable tasks | **~77%** |
| Human-*difficult* tasks | **~23.5%** |

Models solved some problems expert panels could not.

<div style="font-size: 0.5em; color: #5f6b7a; margin-top: 0.8em;">
Anthropic (2026), *Evaluating Claude's bioinformatics research capabilities
with BioMysteryBench*. Problems: <https://huggingface.co/datasets/Anthropic/BioMysteryBench-preview> (CC-BY-4.0)
</div>

::: {.notes}
Accuracy figures are Anthropic's April 2026 release, measured on the original
99-problem set (76 solvable / 23 hard). PLAN.md attributes them to Opus 4.6;
some secondary coverage attributes the same two numbers to Sonnet 4.6 - the
source page presents them as a graph, so CHECK THIS before delivery.

The benchmark was revised in July 2026 (v11): 99 -> 90 problems, 73 solvable
/ 17 hard. Nine were withdrawn after a June audit found the answer keys wrong
or the question underivable from the data provided. That is worth a sentence
out loud - the benchmark's own ground truth needed checking too.

See PLAN.md 3.5.
:::

## What the questions actually look like

::: {.columns}
::: {.column width="50%"}
**Mostly gets right** — solves ~77%

<div style="font-size: 0.72em; line-height: 1.25;">
*"What organism does this crystal structure belong to?"* → `Homo sapiens`

*"What bacteria is found in this sequenced dataset?"* → `Bacillus licheniformis`

*"Identify the transcription factor whose binding sites these peaks
represent."* → `CTCF`
</div>
:::
::: {.column width="50%"}
**Mostly gets wrong** — solves ~23.5%

<div style="font-size: 0.72em; line-height: 1.25;">
*"How many tissue groups are collected, and from how many different
species?"* (metagenomes, non-human hosts) → 2 tissues, 3 snake species

*"What is the most common microRNA associated with the genes given in this
data?"* → any `hsa-let-7` family member
</div>
:::
:::


<div style="font-size: 0.45em; color: #5f6b7a;">
Questions verbatim from the public 5-problem preview, v11 (July 2026); answers
are the published rubrics. Which individual questions Claude failed is not
published — only the per-split rates.
</div>

::: {.notes}
The point of this slide is the shape of the questions, not the answers. Every
one is "here are the files, tell me what they are" - no method is specified,
so the model has to choose the whole pipeline. That is exactly the position
you are in this afternoon.

Note what separates the two columns. The left ones have a single determinate
answer reachable by one canonical route: BLAST it, motif-scan it. The right
ones need several correct choices chained together, and each has a count or a
family in the answer - more places to be almost right. This is the same gap
you will see in the pre-run: it does the standard move well and the judgement
calls invisibly.

Be precise if asked: the right-hand column is the *class* it mostly fails,
not two questions it is known to have failed. Anthropic publishes the split
rates, not per-question results, so nobody can say which individual ones it
got wrong. Worth saying out loud - it is the same "check what is actually
being claimed" move the whole lecture is about.

Do not spend the 1-2-of-5 brittleness finding here - it is the whole of
"Would it do the same thing twice?" at the end of Part 4.

If asked why snakes: the data is anonymised, so the model has to identify
three host species from metagenomic reads without looking up the accession.
:::

## What's AI agents' problem-solving style?

Chen, Zhao & Cohan (2026) — *Measuring the Gap Between Human and LLM
Research Ideas*, arXiv:2607.01233

Took problems from **11,683 papers** and compared of human vs LLM ideas.



## 

| | Human | LLM |
|---|---|---|
| 'Bridge-like' opportunities (connect A to B) | **12.1%** | **47–64%** |
| Synthesis / unification methods | **5.1%** | **22–39%** |
| Diversity | **0.926** | **0.737** |

::: {.fragment}
AI likes synthesising 
:::

## Why this difference?

LLM training: synthesis / average of a huge corpus.

A LLM doesn't have a body, doesn't have a *perspective*: lacks diversity — how would AI evolve?

Overly conservative: an AI reviewer would reject most published papers.





# Part 2 Single cell RNA-seq

##

Simplified central dogma: DNA → **RNA** → protein

- DNA is (nearly) the same in every cell

  - scDNA-seq is niche, e.g. evolution of cancer clones

- scProt-seq is hard to scale: can measure many cells, or many proteins, but not both

- RNA is more transient, but reflects cell activity well




## How it works

1. Dissociate the tissue into single cells
2. Capture each cell in a droplet with a barcoded bead
3. Every transcript from that cell gets the same barcode
4. Sequence everything; sort by barcode

::: {.fragment}
**Output: a genes × cells matrix of counts.**
:::

## Cells or nuclei?

::: {.columns}
::: {.column width="50%"}
**scRNA-seq**

- whole cells
- needs tissue that dissociates gently
- cytoplasmic RNA included
:::
::: {.column width="50%"}
**snRNA-seq**

- nuclei only
- works on frozen / tough tissue
- **cytoplasm discarded**
:::
:::

::: {.fragment}
Does cytoplasmic RNA matter? 

Yes. Eg 'Single-Nucleus RNA-Seq Is Not Suitable for Detection of Microglial Activation Genes in Humans' (Thrupp et al., 2020; Cell Reports)
:::

## Today's dataset

**Sim et al. (2021), *Circulation*** — human heart maturation

| Group | Donors | Age |
|---|---|---|
| foetal | f1–f3 | 19–20 weeks |
| young | y1–y3 | 4–14 years |
| adult | a1–a3 | 35–42 years |


## The matrix

```{r matrix-shape}
seu[["RNA"]]$counts[1:8,1:20]
cat("genes:", nrow(seu), "\n")
cat("nuclei:", ncol(seu), "\n")
m <- SeuratObject::LayerData(seu, assay = "RNA", layer = "counts")
cat("zeros:", round(100 * (1 - length(m@x) / (nrow(m) * ncol(m))), 1), "%\n")
```

<br>

High-dimensional, and **overwhelmingly zero**. Some of those zeros are
'not expressed'; some are 'expressed but not captured'; cannot tell which.

# Part 3 Pre-processing

## Three stages

**Raw data processing** → align reads, produce the count matrix

**Pre-processing** → QC, normalise, select features, correct batch, reduce
dimensions

**Downstream analysis** → cluster, annotate, test hypotheses

<br>

Visualisation throughout. We start at stage 2 — the count matrix is given.

::: {.notes}
Luecken & Theis 2019 figure. Reuse from the 2025 deck.
:::

## Quality control

Not every droplet (unique barcode) is a good cell:

- **Empty droplets** — ambient RNA only; low counts, low gene number
- **Dying cells** — membrane compromised, cytoplasmic RNA leaks out,
  mitochondrial fraction rises
- **Doublets** — two cells, one barcode (hardest to detect)


## QC metrics

```{r qc-violin, fig.width=11, fig.height=4.5}
# Built panel-by-panel rather than with ncol=, so the two count metrics can be
# log-scaled and percent.mt zoomed independently. coord_cartesian, not ylim():
# ylim() would drop the ~1% of nuclei above 20% MT and redraw the violins
# without them.
qc <- VlnPlot(seu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
              group.by = "sample", pt.size = 0, combine = FALSE)
qc[[1]] <- qc[[1]] + scale_y_log10()
qc[[2]] <- qc[[2]] + scale_y_log10()
qc[[3]] <- qc[[3]] + coord_cartesian(ylim = c(0, 20))
wrap_plots(qc, ncol = 3) &
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

Sample level comparison of total RNA counts, genes detected, percentage of **mitochondrial genes**


## How would Claude process the data?

Uploaded the data with **no prompts**

<br>

100 minutes later: QC, integration, clustering, annotation,
composition, differential expression, four figures, four tables, a processed
data object and a written report.

::: {.fragment}
Cost: **$7.09**.
:::

::: {.fragment}
Mostly right decisions, some mistakes.
:::

::: {.notes}
Full scoring in prerun/SCORING_run_A0.md. Every quote from it in this deck is
verbatim. They are spread across the morning rather than shown together: this
slide and the next two here at QC, then one after Integration and one after
Resolution.

This is the first time the pre-run is mentioned all morning. Part 1 was about
AI in science generally - do not pre-empt this there.
:::

## 

Openning of the report:

> "mitochondrial fractions are uniformly low (donor medians 0.10–1.78%) **and**
> so are ribosomal fractions (0.4–1.3%). That combination points to a
> **single-nucleus prep** rather than whole cells, which changes how you'd handle
> ambient RNA."

::: {.fragment}
Nobody told it this was snRNA-seq. **It worked that out from the data** — from
the metric you are looking at — and set its mitochondrial threshold accordingly.
:::



## Normalisation

Cells differ in total counts for both biological and technical reasons, eg sequencing depth (budget) & transcriptional activity.

**Naive log-normalisation**: divide raw counts by lib size, scale, log. 

```{r lognorm-toy, echo=TRUE}
counts <- cbind(A = c(10, 5, 1), B = c(20, 10, 2), C = c(2, 10, 20))
rownames(counts) <- c("ACTB", "GAPDH", "MYH7")
rbind(counts, libsize = colSums(counts))
```

::: {.fragment}
```{r lognorm-toy-2, echo=TRUE}
log1p(t(t(counts) / colSums(counts)) * 1e4)
```
:::

::: {.fragment}
**A and B are now identical**
:::


## More sophisticated methods

- **scran**: pool cells (clustering) first to estimate cluster-wise lib size (more robust).
- **SCTransform**: negative binomial regression: regress lib size out, use residual as normalised values.

::: {.fragment}
These methods don't stricly seperate bio vs tech variations in lib size. You need spike-ins (synthetic negative control RNAs).
:::



## Feature selection

~19,000 genes. Most are uninformative:

- not expressed
- expressed everywhere at the same level (housekeeping)

<br>

Keep the **2,000–5,000 most variable** genes. 

::: {.fragment}
Caveat: a gene you filter out cannot be discovered.
:::

## Dimension reduction: PCA

Still thousands of dimensions, and they are correlated.

**PCA** finds linear combinations capturing the most variance.

- Use the first 10–50 PCs for everything downstream
- Faster, and drops noise in the tail components
- **Not good** for visualisation — 2 PCs is not enough


## Visualisation: UMAP

Non-linear projections of high-dim data (or their PCs) to 2D. Good at preserving **local** structure: cells that were neighbours stay neighbours.

UMAP starts with kNN graph:

```{r knn-toy, fig.width=10, fig.height=3.8}
p1 <- toy_base(ggplot(toy, aes(x, y)) + geom_point(size = 3, colour = "grey30"),
               paste(nrow(toy), "cells in 20-dimensional PC space"))
p2 <- toy_base(ggplot(toy, aes(x, y)) + toy_segs() +
               geom_point(size = 3, colour = "grey30"),
               paste0("join each to its k = ", K_toy, " nearest"))
p1 + p2
```

Throw away the coordinates. Keep only **who is near whom**.

::: {.notes}
Say explicitly that the left panel is a cartoon: real cells live in 20 PCs, and
the whole problem is that we cannot see that. Everything from here on uses only
the edges, never the positions.

Ask how many groups they see before advancing. The right-hand pair is meant to
be arguable - it is the pair that merges on the resolution slide later.
:::

## How UMAP works

::: {.incremental}
1. Build that **k-nearest-neighbour graph** in the high-dimensional space of RNA or PC.
2. Weight each edge by distance — but rescaled **per cell**, so every cell gets
   roughly the same total connection, dense regions and sparse regions alike
3. Lay the graph out in 2D: connected cells **attract**, unconnected cells
   **repel**, until it settles
:::

<br>

::: {.fragment}
Step 2 throws away local **density** and **global** distance.

The gains: rare cell type more visible; clusters separable
:::



## On our data

```{r umap-demo, fig.width=9, fig.height=4}
s2 <- readRDS(file.path(D, "checkpoints/02_integrated_clustered.rds"))
DimPlot(s2, reduction = "umap", group.by = "group", cols = group_cols,
        shuffle = TRUE) + ggtitle("UMAP, coloured by developmental stage")
```

## UMAP: what it does not mean {background-color="#F5E9E9"}

::: {.incremental}
- **Distance between clusters is not meaningful**
- **Cluster area is not population size**
- Apparent gaps can be artefacts of the parameters
- Clusters can appear that are not there
:::

<br>

::: {.fragment}
It is a picture for looking at possible clusters. Never test a hypothesis on UMAP coordinates.
:::


## Batch effects

```{r batch, fig.width=10, fig.height=4.2}
# UMAP on the *unintegrated* PCA, so both panels are the same kind of picture
s2 <- RunUMAP(s2, reduction = "pca", dims = 1:20,
              reduction.name = "umap_unint", verbose = FALSE)
p1 <- DimPlot(s2, reduction = "umap_unint", group.by = "sample", shuffle = TRUE) +
    ggtitle("UMAP — before integration") + theme(legend.position = "none")
p2 <- DimPlot(s2, reduction = "umap", group.by = "sample", shuffle = TRUE) +
    ggtitle("UMAP — after Harmony")
p1 + p2
```

Before integration, cells clustered by **donor**: is it good or bad?


## A design problem

Batch effects are most effectively dealt with at experimental design stage:

::: {.incremental}
- Case 1: Treated and control cells from the *same* donor, hashed and run in
  **one** 10x lane. 
- Case 2: Six donors, three conditions, run over four days —
  with **each day carrying a mix of conditions**. 
- Case 3: All patients in January on chemistry v2, all controls in June
  on v3. Batch and biology are the same column.
:::




## How Claude Sci evaluated batches


> "Clustering on unintegrated PCA gave clusters that were stage-exclusive *and*
> donor-exclusive — 7 clusters were ≥90% a single donor. That can't distinguish
> development from batch. After Harmony: **0** donor-dominated clusters, while
> **10** clusters stay stage-restricted. Since those now draw from all donors
> within a stage, the restriction is developmental."

::: {.fragment}
Importance of biological repeats.
:::



# Part 4 Clustering

## Graph-based clustering

1. Build a **shared nearest neighbour** graph — cells connected by how many
   neighbours they share
2. Find **communities** in that graph

<br>

**Louvain** (Seurat default) · **Leiden** (faster, better-connected communities)

## Why *shared* neighbours?

Starting with same kNN graph for UMAP. But re-weight every edge by how
many neighbours its two endpoints have in common.

::: {.incremental}
- Two cells of the same type in the same crowd: they share **most** of
  their k neighbours 
- Two cells that happen to be adjacent across a boundary have very different kNNs → the edge survives kNN but is nearly weightless
:::



## 

sNN graph for toy data with 4 clusters 

```{r comm-toy, fig.width=6.5, fig.height=4.3}
toy_base(ggplot(toy, aes(x, y)) + toy_segs(TRUE) +
    scale_linewidth(range = c(.15, 1.6)) + scale_alpha(range = c(.1, .85)) +
    geom_point(aes(colour = toy_cl(0.5)), size = 3) +
    scale_colour_manual(values = toy_pal),
    "line width = shared neighbours")
```

```{r snn-numbers}
cl5 <- toy_cl(0.5)
btw <- max(ed$shared[cl5[ed$i] != cl5[ed$j]])
```



## Resolution

```{r res-toy, fig.width=12, fig.height=3.7}
wrap_plots(lapply(c(0.1, 0.5, 2), function(r) toy_base(
    ggplot(toy, aes(x, y)) + toy_segs(TRUE) +
    scale_linewidth(range = c(.15, 1.4)) + scale_alpha(range = c(.08, .7)) +
    geom_point(aes(colour = toy_cl(r)), size = 2.8) +
    scale_colour_manual(values = toy_pal),
    paste0("resolution = ", r, "  →  ", nlevels(toy_cl(r)), " clusters"))),
    nrow = 1)
```

Resolution parameter: how much internal density a group needs to form its own cluster.


## Resolution on our heart data

```{r res-counts}
res <- c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8)
n <- sapply(res, function(r) length(unique(s2@meta.data[[paste0("SCT_snn_res.", r)]])))
knitr::kable(data.frame(resolution = res, clusters = n))
```

::: {.fragment}
How many clusters?
:::


## What's left for us to do?

From Anthropic's own evaluation, on the hard problems:

> nearly half of the wins are paths the model *stumbles onto* rather than
> reproduces

In my first run, Claude Sci didn't realise the data was snRNA-seq rather than sc.

::: {.callout-note appearance="minimal"}

## Practical 1 
**Decision 1.** Where do you set your QC thresholds?

**Decision 2.** How many PCs?

**Decision 3.** How many clusters are real?
:::









