---
title: "Single-cell RNA-seq"
subtitle: "From counts to clusters · BINF90004 Bioinformatics Case Studies"
author: "Dr Jiadong Mao · Melbourne Integrative Genomics"
date: today
format:
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true          # press 'b' to draw, 'c' for chalkboard
    incremental: false
    width: 1280
    height: 720
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
suppressPackageStartupMessages({
    library(Seurat); library(ggplot2); library(dplyr); library(patchwork)
})
D <- if (dir.exists("data")) "data" else "../data"
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")

# Loaded here rather than mid-deck: two later slides use it, and if it is
# loaded inside a slide chunk, deleting that slide silently breaks them.
seu <- readRDS(file.path(D, "P1_start.rds"))
```

# Part 1 {.center}

## What does a bioinformatician do now? {.center}

::: {.notes}
12 minutes. This is not a technology showcase - it is the reason the practical
is built the way it is. Do not rush it.
:::

## A show of hands

**Who has used an LLM to write analysis code in the last month?**

<br>

**Who checked what it chose for you?**

::: {.notes}
Expect nearly every hand for the first, very few for the second. That gap is
the whole lecture.
:::

## This is a product category now

::: {.incremental}
- **Claude Science** (beta, 2026) — multi-agent workbench for scientific analysis
- Connectors: Benchling, 10x Genomics, PubMed, BioRender, Synapse
- **Agent Skills** — packaged, reproducible analysis workflows
:::

<br>

::: {.fragment}
The first Agent Skill Anthropic released is **`single-cell-rna-qc`**.

It automates the quality control step of scRNA-seq.

**That is section 2 of your practical this afternoon.**
:::

## It is genuinely good

**BioMysteryBench** — 99 expert-written questions on real datasets, objective
ground truth

| | Accuracy |
|---|---|
| Human-solvable tasks (76) | **~77%** |
| Human-*difficult* tasks (23) | **~23.5%** |

Models solved some problems expert panels could not.

::: {.fragment}
One of the actual tasks: *"identify which human organ this scRNA-seq dataset
came from."*
:::

::: {.notes}
Re-check these numbers before delivery - model versions move. See PLAN.md 3.5.
:::

## So I gave it our dataset

I uploaded the folder you are about to analyse. **No question, no instructions**
— just the files.

<br>

100 minutes later, unattended: QC, integration, clustering, annotation,
composition, differential expression, four figures, four tables, a processed
data object and a written report.

::: {.fragment}
Cost: **$7.09**.
:::

::: {.notes}
Full scoring in prerun/SCORING_run_A0.md. Everything on the next slides is
quoted verbatim from its report.
:::

## The first thing it said {.center}

> "mitochondrial fractions are uniformly low (donor medians 0.10–1.78%) **and**
> so are ribosomal fractions (0.4–1.3%). That combination points to a
> **single-nucleus prep** rather than whole cells, which changes how you'd handle
> ambient RNA."

::: {.fragment}
Nobody told it this was snRNA-seq. **It worked that out from the data**, and set
its mitochondrial threshold accordingly.
:::

## I had predicted it would get that wrong

I expected `percent.mt < 5` — the whole-cell default from every tutorial.

::: {.fragment}
<br>

I had also written down the "expert" rebuttal:

> *Heart is mitochondria-rich. Cardiomyocytes are packed with mitochondria. An
> aggressive MT filter deletes the cell type the study is about.*
:::

::: {.fragment}
<br>

**That reasoning is also wrong.** Median `percent.mt` here is 0.25%, and
cardiomyocytes have the *lowest* of any cell type — because these are nuclei and
the cytoplasm has been washed away.

**Its reasoning beat mine.**
:::

::: {.notes}
Own this. Saying "I got this wrong and the machine didn't" is the most
persuasive thing you can do in this room, and it is true.
:::

## It also did something our practical does not

On whether integration destroyed the biology:

> "Clustering on unintegrated PCA gave clusters that were stage-exclusive *and*
> donor-exclusive — 7 clusters were ≥90% a single donor. That can't distinguish
> development from batch. After Harmony: **0** donor-dominated clusters, while
> **10** clusters stay stage-restricted. Since those now draw from all donors
> within a stage, the restriction is developmental."

::: {.fragment}
It designed a test to tell correct integration from over-correction. Your
practical only warns you that the difference exists.
:::

## And then, in its caveats {.center}

> "Wilcoxon on individual cells treats cells as replicates and **inflates
> significance** — for anything you'd report, redo this as donor-level pseudobulk
> with DESeq2 at n=3 per stage."

::: {.fragment}
That is completely correct. We will spend twenty minutes on exactly this point
after lunch.
:::

## But this is its headline {background-color="#F5E9E9"}

> "Adult vs. fetal cardiomyocytes: **5,967 significant genes**"

<br>

With a volcano plot. Pseudobulk appears in the deliverable only as a
*heatmap* — it was never used as a test.

::: {.fragment}
<br>

**It knew the number was wrong. It led with it anyway.**
:::

## The failure is not ignorance {.center}

> The caveat and the number do not carry equal weight.
>
> A reader who skims takes away 5,967. Only someone who **already knows what to
> look for** reads to the end and discounts it.

::: {.notes}
This is the thesis of the day. Say it slowly.
:::

## And the one it did not mention

It reports cardiomyocytes falling from **64.4%** of cells to **23.3%** and calls
the change biological.

::: {.fragment}
It ran **no statistical test of composition at all.** No propeller, no scCODA, no
chi-square. Nothing.
:::

::: {.fragment}
<br>

And unlike the differential expression problem, **that limitation appears nowhere
in its caveats.**
:::

## The asymmetry {.center}

> It flagged the error it had a name for.
>
> It was silent about the one it didn't.
>
> **Which is exactly the failure you cannot catch by reading the caveats.**

## But what about taste?

**Chen, Zhao & Cohan (2026)** — *Measuring the Gap Between Human and LLM
Research Ideas*, arXiv:2607.01233

They do not score ideas one at a time. They reconstruct the literature context
for **11,683 papers** and compare the *distribution* of human vs LLM ideas.

::: {.fragment}
**Any single LLM idea looks fine. The narrowness only shows up in aggregate.**
:::

## The gap

| | Human | LLM |
|---|---|---|
| "Bridge-like" opportunities (connect A to B) | **12.1%** | **47–64%** |
| Synthesis / unification methods | **5.1%** | **22–39%** |
| Diversity (normalised entropy) | **0.926** | **0.737** |

::: {.fragment}
And extended reasoning made it **worse** — one model went from 49.7% → **71.1%**
bridge-like with thinking mode on.
:::

## Why you should care

LLM research ideas cluster on:

- *"combine X with Y"*
- *"apply method A to dataset B"*

<br>

::: {.fragment}
*"Apply a foundation model to spatial transcriptomics."*

*"Integrate multi-omics with a graph neural network."*

<br>

Recognise those? Some of you have written them in a project proposal.
:::

::: {.notes}
This is the moment the room goes quiet. Let it.
Be fair to the paper: STEM-only corpus, one-shot prompting, reconstructed
contexts lack tacit knowledge. The claim is "current LLMs, prompted this way,
produce a narrow distribution" - NOT "AI cannot have taste."
:::

## Would it do the same thing twice?

From Anthropic's own evaluation, on the hard problems:

> nearly half of the wins are paths the model *stumbles onto* rather than
> reproduces

::: {.fragment}
<br>

**I have run this once.** So I cannot tell you whether a second run picks the
same thresholds, the same resolution, or notices the nuclei at all.

Neither could you, from a paper that reports one run.
:::

::: {.notes}
HONEST STATUS: n = 1. Do NOT claim run-to-run variance until A1-A3 are done -
see prerun/SCORING_run_A0.md. If you do get the repeat runs before the session,
replace this slide with the actual comparison; it is much stronger.
:::

## So what is left for you?

::: {.incremental}
- Not knowing more statistics than it does. It diagnosed pseudoreplication
  unprompted.
- Not knowing more biology. It identified the assay from the data.
- **Reading to the end.** Noticing which number went in the headline.
- **Noticing the test that was never run** — and never mentioned.
:::

::: {.fragment}
<br>

Every one of those requires you to already understand the method well enough to
know what *should* have been there.
:::

## The deal for today

You will make **five decisions** this afternoon that no tool makes for you:

1. Where to set your quality thresholds
2. How many principal components to keep
3. How finely to cluster
4. What cell type a cluster is
5. How to test for differential expression

::: {.fragment}
Each time: choose, then **justify it to the person next to you**.

That is the assessable skill, and it is what your report is marked on.
:::

# Part 2 {.center}

## Why single cell {.center}

## Why RNA

DNA → **RNA** → protein

- DNA is the same in (nearly) every cell
- Protein is hard to measure at scale
- RNA is what the cell is *doing right now*, and we can count it

## Bulk vs single cell

**Bulk RNA-seq**: average over millions of cells

::: {.fragment}
A tissue that is 70% cardiomyocytes and 5% immune cells gives you a
cardiomyocyte-flavoured average, and the immune biology is invisible.
:::

::: {.fragment}
**scRNA-seq**: one measurement per cell. You can ask *which* cells, not just
*how much*.
:::

## How it works

1. Dissociate the tissue into single cells
2. Capture each cell in a droplet with a barcoded bead
3. Every transcript from that cell gets the same barcode
4. Sequence everything; sort by barcode

::: {.fragment}
**Output: a genes × cells matrix of counts.**
:::

## Cells or nuclei?

::: {.columns}
::: {.column width="50%"}
**scRNA-seq**

- whole cells
- needs tissue that dissociates gently
- cytoplasmic RNA included
:::
::: {.column width="50%"}
**snRNA-seq**

- nuclei only
- works on frozen / tough tissue
- **cytoplasm discarded**
:::
:::

::: {.fragment}
Adult heart muscle does not dissociate nicely. **Today's data is snRNA-seq** —
which is why `percent.mt` behaved the way it did ten minutes ago.
:::

## The matrix

```{r matrix-shape}
cat("genes:", nrow(seu), "\n")
cat("nuclei:", ncol(seu), "\n")
m <- SeuratObject::LayerData(seu, assay = "RNA", layer = "counts")
cat("zeros:", round(100 * (1 - length(m@x) / (nrow(m) * ncol(m))), 1), "%\n")
```

<br>

High-dimensional, and **overwhelmingly zero**. Some of those zeros are
"not expressed"; some are "expressed but not captured". You cannot tell which.

# Part 3 {.center}

## The workflow {.center}

## Three stages

**Raw data processing** → align reads, produce the count matrix

**Pre-processing** → QC, normalise, select features, correct batch, reduce
dimensions

**Downstream analysis** → cluster, annotate, test hypotheses

<br>

Visualisation throughout. We start at stage 2 — the count matrix is given.

::: {.notes}
Luecken & Theis 2019 figure. Reuse from the 2025 deck.
:::

## Today's dataset

**Sim et al. (2021), *Circulation*** — human heart maturation

| Group | Donors | Age |
|---|---|---|
| foetal | f1–f3 | 19–20 weeks |
| young | y1–y3 | 4–14 years |
| adult | a1–a3 | 35–42 years |

::: {.fragment}
**Nine donors.** Remember that number — in the second lecture it turns out to be
the only number that matters.
:::

# Part 4 {.center}

## Pre-processing {.center}

## Quality control

Not every barcode is a good cell:

- **Empty droplets** — ambient RNA only; low counts, low gene number
- **Dying cells** — membrane compromised, cytoplasmic RNA leaks out,
  mitochondrial fraction rises
- **Doublets** — two cells, one barcode; look like a chimera

::: {.fragment}
Tools exist for each (CellBender, scDblFinder, DoubletFinder). None of them
removes the need to choose.
:::

## QC metrics

```{r qc-violin, fig.width=11, fig.height=4.5}
VlnPlot(seu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
        group.by = "sample", pt.size = 0, ncol = 3) &
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

**Look at the donors, not just the distribution.** They differ, and a single
global threshold lands differently on each.

## Normalisation

Cells differ in total counts for reasons that are not biological — capture
efficiency, sequencing depth, cell size.

::: {.incremental}
- **Log-normalisation**: divide by library size, scale, log. Simple, fast.
- **SCTransform**: regularised negative binomial; returns Pearson residuals.
  Handles the mean–variance relationship properly.
:::

::: {.fragment}
We use SCTransform today. Neither is "correct" — they answer slightly different
questions.
:::

## Feature selection

~19,000 genes. Most are uninformative:

- not expressed
- expressed everywhere at the same level (housekeeping)

<br>

Keep the **2,000–5,000 most variable** genes. Everything downstream — PCA,
clustering — runs on those.

::: {.fragment}
Note what this means: **a gene you filter out cannot be discovered.**
:::

## Dimension reduction: PCA

Still thousands of dimensions, and they are correlated.

**PCA** finds linear combinations capturing the most variance.

- Use the first 10–50 PCs for everything downstream
- Faster, and drops noise in the high components
- **Not** for visualisation — 2 PCs is not enough

::: {.fragment}
How many to keep? An elbow plot, and a judgement call. *(Decision 2.)*
:::

## Visualisation: UMAP and t-SNE

Non-linear projections to 2D. Good at preserving **local** structure — cells
that were neighbours stay neighbours.

```{r umap-demo, fig.width=9, fig.height=4}
s2 <- readRDS(file.path(D, "checkpoints/02_integrated_clustered.rds"))
DimPlot(s2, reduction = "umap", group.by = "group", cols = group_cols,
        shuffle = TRUE) + ggtitle("UMAP, coloured by developmental stage")
```

## UMAP: what it does not mean {background-color="#F5E9E9"}

::: {.incremental}
- **Distance between clusters is not meaningful**
- **Cluster area is not population size**
- Apparent gaps can be artefacts of the parameters
- Clusters can appear that are not there
:::

<br>

::: {.fragment}
It is a picture for looking at, not a measurement. Never test a hypothesis on
UMAP coordinates.
:::

## Batch effects

```{r batch, fig.width=10, fig.height=4.2}
p1 <- DimPlot(s2, reduction = "pca", group.by = "sample", shuffle = TRUE) +
    ggtitle("PCA — before integration") + theme(legend.position = "none")
p2 <- DimPlot(s2, reduction = "harmony", group.by = "sample", shuffle = TRUE) +
    ggtitle("After Harmony")
p1 + p2
```

Cells cluster by **donor** instead of by cell type. Each sample carries its own
technical signature.

## Integration

**MNN** · **CCA** · **Harmony** — all find a shared space where donors overlap.

::: {.fragment}
::: {.callout-warning appearance="minimal"}
**There is no measurement that tells you it worked.**

"The donors mix better" is also exactly what over-correction looks like — and
over-correction deletes real biology.
:::
:::

::: {.fragment}
Here the donors *differ by developmental stage*. Some separation between them is
the signal we came to study. We integrate by **donor**, not by **stage** —
deliberately.
:::

# Part 5 {.center}

## Clustering {.center}

## Graph-based clustering

1. Build a **shared nearest neighbour** graph — cells connected by how many
   neighbours they share
2. Find **communities** in that graph

<br>

**Louvain** (Seurat default) · **Leiden** (faster, better-connected communities)

## Resolution

```{r res-counts}
res <- c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8)
n <- sapply(res, function(r) length(unique(s2@meta.data[[paste0("SCT_snn_res.", r)]])))
knitr::kable(data.frame(resolution = res, clusters = n))
```

::: {.fragment}
There is no correct value. It depends on the question — and today, on you.
*(Decision 3.)*
:::

## A warning you will verify this afternoon

The published analysis of this tissue describes **8 cell types**.

::: {.fragment}
At the *lowest* resolution above, you already get **12 clusters**.
:::

::: {.fragment}
<br>

**A cluster is not a cell type.** It is a community in a graph you built, using
genes you selected, in a space you chose the dimension of.
:::

## What you will do now

::: {.callout-note appearance="minimal"}
## Practical 1 — 65 minutes
**Decision 1.** Where do you set your QC thresholds?

**Decision 2.** How many PCs?

**Decision 3.** How many clusters are real?
:::

<br>

There is one hard sync point. You are not expected to finish every chunk.

If your R is broken, you can still do all three — **come anyway.**
