---
title: "Single-cell RNA-seq"
subtitle: "From clusters to conclusions · BINF90004 Bioinformatics Case Studies"
author: "Dr Jiadong Mao · Melbourne Integrative Genomics"
date: today
format:
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true
    width: 1280
    height: 720
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
suppressPackageStartupMessages({
    library(Seurat); library(ggplot2); library(dplyr); library(speckle)
})
D <- if (dir.exists("data")) "data" else "../data"
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")
s2  <- readRDS(file.path(D, "checkpoints/02_integrated_clustered.rds"))
key <- readRDS(file.path(D, "celltype_key.rds"))
lab <- readRDS(file.path(D, "cluster_labels.rds"))
s2$celltype <- unname(lab[as.character(s2$seurat_clusters)])
```

## Where we are

You have **19 clusters**. Clusters are numbered blobs.

<br>

Today's second half: give them names, then use those names to make two
biological claims.

::: {.fragment}
1. Which cell types change in abundance as the heart matures?
2. Which genes change *within* a cell type?
:::

::: {.fragment}
The second is where most published single-cell analyses go wrong.
:::

# Part 1 {.center}

## Annotation {.center}

## Three ways to name a cluster

::: {.incremental}
**Manual, marker-based** — find genes enriched in the cluster, look them up.
Flexible, slow, needs domain knowledge.

**Reference-based** — map onto an annotated atlas (Azimuth, SingleR,
CellTypist). Fast, reproducible, fails on cell types absent from the reference.

**Soft / continuous** — score every cell against every reference phenotype
instead of assigning one label.
:::

## Why hard labels are uncomfortable

A cell is not obliged to belong to a category we invented.

- Differentiating cells are genuinely between states
- Doublets are genuinely two things
- Novel cell types are genuinely not in your reference

::: {.fragment}
**Φ-Space** (our lab) replaces the label with a vector of scores — one per
reference phenotype. Cells that sit between types *look* like they sit between
types.
:::

::: {.notes}
Pull figures from afternoon_30min_xaringan.Rmd in the workshop repo.
Keep this brief - 2 slides. It is a signpost, not a tutorial.
:::

## The trap

```{r purity}
tb <- table(s2$seurat_clusters, key[colnames(s2)])
df <- data.frame(
    cluster = rownames(tb),
    n = as.integer(rowSums(tb)),
    published = colnames(tb)[apply(tb, 1, which.max)],
    agreement = paste0(round(100 * apply(tb, 1, max) / rowSums(tb)), "%"))
knitr::kable(head(df[order(-df$n), ], 10))
```

## Two failures at once {background-color="#F5E9E9"}

**Too fine.** 19 clusters, and **11 of them are cardiomyocytes**.

::: {.fragment}
**Too coarse.** The published analysis has 8 cell types. You recover **6**.
Smooth muscle (62 nuclei) and erythroid (9) get no cluster at all — 60 of the 62
smooth muscle nuclei are hidden inside cluster 3.
:::

::: {.fragment}
<br>

**Nothing in the output flagged either one.** Every one of those 11
cardiomyocyte clusters has perfectly good marker genes.
:::

## Rare populations

Two independent ways to lose them, both of which you have now seen:

1. **QC thresholds** — endothelial and immune cells were hit hardest by
   `percent.mt < 5`
2. **Clustering** — smooth muscle never formed a cluster of its own

<br>

::: {.fragment}
Both failures point the same direction. If your question is about a rare cell
type, every default in the pipeline is working against you.
:::

# Part 2 {.center}

## Composition {.center}

## Do proportions change?

```{r comp, fig.width=10, fig.height=4.5}
s2@meta.data |>
    count(sample, group, celltype) |>
    group_by(sample) |> mutate(prop = n / sum(n)) |> ungroup() |>
    ggplot(aes(sample, prop, fill = celltype)) +
    geom_col() +
    facet_grid(~ group, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "proportion of nuclei", fill = NULL) +
    theme_minimal()
```

## The wrong way

A chi-square test on cell counts.

::: {.fragment}
With ~7,000 nuclei, almost everything is significant.
:::

::: {.fragment}
<br>

**Your independent units are the nine donors, not the nuclei.** Nuclei from one
donor are subsamples of a single biological observation.
:::

## propeller

```{r propeller}
pr <- propeller(clusters = s2$celltype, sample = s2$sample, group = s2$group)
knitr::kable(pr[, c("PropMean.fetal", "PropMean.young", "PropMean.adult", "FDR")],
             digits = 4)
```

Computes each **donor's** proportions, transforms them, tests across donors.

## What it says

Cardiomyocytes fall from **~74%** of nuclei in foetal hearts to **~24%** in
adult. Immune cells and fibroblasts rise.

<br>

::: {.fragment}
Two things before you believe it:

1. These are **nuclear** proportions. Adult cardiomyocytes are binucleated and
   polyploid — part of this is the denominator changing, not the tissue.
2. Compositions **sum to 1**. If cardiomyocytes fall, everything else rises
   whether or not anything happened to it.
:::

# Part 3 {.center}

## Differential expression {.center}

## The question

Within cardiomyocytes: which genes differ between adult and foetal hearts?

<br>

::: {.fragment}
Every tutorial reaches for `FindMarkers()`. On this data it returns
**~2,000 significant genes**.
:::

::: {.fragment}
That number is wrong, and the reason is the most important thing in this
lecture.
:::

## Pseudoreplication

`FindMarkers` treats **every nucleus as an independent replicate**.

They are not. Nuclei from one donor share:

- that donor's genome
- their age, sex, medication, disease history
- one tissue dissociation, one library prep, one sequencing run

::: {.fragment}
<br>

**Effective sample size is 3 donors per group, not 3,200 nuclei.**
:::

## A control that should find nothing

Take two donors at the **same** developmental stage. There is no developmental
difference to find.

```{r demo, fig.width=8, fig.height=3.6}
demo <- readRDS(file.path(D, "de_demo.rds"))
ggplot(demo$wilcoxon, aes(reorder(comparison, n_sig), n_sig, fill = kind)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c(`real contrast` = "#2980B9",
                                 `same stage (foetal)` = "#C0392B")) +
    labs(x = NULL, y = "genes called significant", fill = NULL,
         subtitle = "per-nucleus Wilcoxon, 300 cells per side") +
    theme_minimal()
```

## Read that again {background-color="#F5E9E9"}

Two foetal donors — same stage, same tissue, same protocol — produce
**hundreds of "significantly differentially expressed" genes**.

<br>

::: {.fragment}
There is no biology in that comparison. Those genes are **donor variation**
reported to you as a finding.
:::

::: {.fragment}
The noise floor is **~25%** of the "real" adult-vs-foetal result.
:::

## What that means

When the per-nucleus test says 2,000 genes separate adult from foetal
cardiomyocytes, it **cannot distinguish**:

<br>

*"adult hearts differ from foetal hearts"*

from

*"these three donors differ from those three donors"*

<br>

::: {.fragment}
It never modelled donor. It has no way to tell them apart.
:::

## Pseudobulk

Sum counts within each **donor × cell type**. One profile per donor.

<br>

Then use methods built for a handful of replicates — which is what bulk RNA-seq
methods are. `edgeR`, `limma-voom`, `DESeq2`.

::: {.fragment}
<br>

Nine columns instead of seven thousand. **That was always the real sample size.**
:::

## And the control?

Run the same two-foetal-donor comparison through pseudobulk:

<br>

::: {.fragment}
It **refuses**. One sample per group — there is no residual variance to
estimate.
:::

::: {.fragment}
<br>

**That refusal is the correct answer.** The per-nucleus test does not refuse. It
manufactures thousands of degrees of freedom from one donor and hands you a gene
list.
:::

::: {.notes}
Squair et al., Nature Communications 2021. Cite it.
:::

## What you can honestly conclude

- n = 3 per group. Small.
- Sex is confounded with stage (2M/1F, 2M/1F, 1M/2F)
- Donors differ in depth and quality — `a3` loses far more nuclei to any QC
  threshold than `f1`
- Cell type labels are an interpretation, not a measurement
- Pseudobulk discards within-type heterogeneity by construction

::: {.fragment}
Your report needs a section like this. It is not a weakness to admit them —
it is the difference between a result and a claim.
:::

# Part 4 {.center}

## Where the field is going {.center}

## Beyond discrete labels

**Trajectory / pseudotime** — order cells along a continuous process rather than
binning them. Slingshot, Monocle3, DPT, RNA velocity.

**Cell–cell interaction** — infer signalling between cell types from
ligand–receptor expression. CellPhoneDB, CellChat.

**Cell-specific networks** — a co-expression network *per cell* rather than per
cluster.

::: {.notes}
One slide each at most, with a figure from the workshop deck. This is a
signpost section - do not teach the methods.
:::

## Spatial transcriptomics

scRNA-seq tells you **which cells are doing what**.

Spatial tells you **where they are doing it**.

::: {.fragment}
- Sequencing-based: Visium, Visium HD, Slide-seq
- Imaging-based: Xenium, CosMx, MERFISH
:::

::: {.fragment}
Tumour microenvironment, tissue architecture, niches — questions dissociation
destroys.
:::

## What is a cell, spatially?

Resolutions: super-cellular, near-cellular, sub-cellular

**The common problem: mixed cell identities**

- imperfect segmentation
- arbitrary binning
- transcript diffusion and misassignment

::: {.fragment}
The same "hard labels are uncomfortable" problem, in a new place.
:::

::: {.notes}
Reuse the two spatial slides from the 2025 deck - they are still good.
:::

## Foundation models

scGPT, Geneformer, scFoundation, UCE — transformers pre-trained on tens of
millions of cells.

::: {.fragment}
Promise: transfer to new tissues, few-shot annotation, in-silico perturbation.
:::

::: {.fragment}
Open question: **do they beat well-tuned simple baselines?** For several tasks
the honest answer is still "not clearly".
:::

::: {.fragment}
A good report topic, and a good place to practise the scepticism from this
morning.
:::

# The five decisions {.center}

## What you chose, and what a default would have chosen

| | Default | You |
|---|---|---|
| QC thresholds | `percent.mt < 5` | ? |
| Number of PCs | 10 | ? |
| Clustering resolution | 0.8 | ? |
| Cluster identity | whatever the markers suggest | ? |
| DE method | per-nucleus Wilcoxon | ? |

::: {.fragment}
**Every one of these ran without an error message either way.**
:::

## What the agent did, scored

| Step | Its choice | OK here? | Said so? |
|---|---|---|---|
| Assay | inferred **snRNA-seq** from MT + ribo | ✅ | ✅ |
| QC | ≥500 genes, ≥1000 UMI, MT <10% | ✅ | ✅ |
| Per-donor loss | flagged `y3`, `a3` | ✅ | ✅ |
| Integration | Harmony **+ explicit batch-vs-biology test** | ✅ | ✅ |
| Clustering | Leiden, res 0.6 | ✅ | ✅ |
| Annotation | 11 canonical categories | ✅ | ✅ |
| **Composition test** | **none** | ❌ | ❌ |
| **DE** | **per-cell Wilcoxon → 5,967** | ❌ | ⚠️ last |

::: {.notes}
Full detail in prerun/SCORING_run_A0.md. 100 min, $7.09, one unprompted run.
:::

## Same data, two analyses

| | Claude Science | You |
|---|---|---|
| Cells after QC | 7,409 | 6,982 |
| CM foetal → adult | 64.4% → 23.3% | 74% → 24% |
| Composition test | **none** | propeller, 5/6 significant |
| CM adult vs foetal | **5,967** (per-cell) | 2,019 per-cell → **1,569 pseudobulk** |

::: {.fragment}
The *data* agree. The disagreement is entirely about **inference** — and that is
the part you were trained for.
:::

## Back to this morning {.center}

> The part of the job that is hardest to automate is not having an idea.
>
> It is having an idea that isn't the same idea everyone else's model just had —
> and being able to tell whether the answer on the screen is real.

## Your report

**TODO — dataset options and brief.**

::: {.notes}
Confirm the assessment brief with the subject coordinator. JM does not mark it.
:::
