---
title: "Single-cell RNA-seq"
subtitle: "Module II: From clusters to conclusions"
author: "Dr Jiadong Mao </br> Melbourne Integrative Genomics (MIG) </br> BINF90004 Bioinformatics Case Studies 2026"
format:
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true          # press 'b' to draw, 'c' for chalkboard
    incremental: false
    width: 1280
    height: 720
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
suppressPackageStartupMessages({
    library(Seurat); library(ggplot2); library(dplyr); library(speckle)
})
# Resolve the data folder for every layout this deck is rendered from:
# repo root, repo slides/, and the private authoring folder (../dist).
D <- if (dir.exists("data")) "data" else
     if (dir.exists("../data")) "../data" else "../dist"
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")
# Loaded here rather than mid-deck: several later slides use these, and if they
# are loaded inside a slide chunk, deleting that slide silently breaks them.
s2  <- readRDS(file.path(D, "checkpoints/02_integrated_clustered.rds"))
key <- readRDS(file.path(D, "celltype_key.rds"))
lab <- readRDS(file.path(D, "cluster_labels.rds"))
s2$celltype <- unname(lab[as.character(s2$seurat_clusters)])
```

## Where we are

You have **19 clusters**. Clusters are numbered blobs.

<br>

Today's second half: give them names, then use those names to make two
biological claims.

1. Which cell types change in abundance as the heart matures?
2. Which genes change *within* a cell type across dev stages?



# Part 1 Annotation

## Three ways to name a cluster

::: {.incremental}
**Manual, marker-based** — find genes enriched in the cluster, look them up.
Flexible, slow, needs domain knowledge.

**Reference-based** — map onto an annotated atlas (Azimuth, SingleR,
CellTypist). Fast, reproducible, fails on cell types absent from the reference.

**Soft / continuous** — score every cell against every reference phenotype
instead of assigning one label. Eg **Φ-Space**
:::


## Where are the rare cells

Possible ways to miss them:

1. **QC thresholds** — endothelial and immune cells were hit hardest by
   `percent.mt < 5`
2. **Clustering** — smooth muscle never formed a cluster of its own



# Part 2 Composition

## Do proportions change?

```{r comp, fig.width=10, fig.height=4.5}
s2@meta.data |>
    count(sample, group, celltype) |>
    group_by(sample) |> mutate(prop = n / sum(n)) |> ungroup() |>
    ggplot(aes(sample, prop, fill = celltype)) +
    geom_col() +
    facet_grid(~ group, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "proportion of nuclei", fill = NULL) +
    theme_minimal()
```


## propeller

Your independent units are the **nine donors**

::: {.incremental}
1. Compute each **donor's** proportion of every cell type 
2. **Transform** them (logit or arcsin-square-root) to Euclidean space
3. Fit a linear model per cell type, **borrowing information across
   cell types** (empirical Bayes machinery, same as limma)
4. **t-test** for two groups; **ANOVA F-test** for more than two
:::

<div style="font-size: 0.5em; color: #5f6b7a; margin-top: 0.8em;">
Phipson, Sim, Porrello, Hewitt, Powell & Oshlack (2022), *propeller: testing for
differences in cell type proportions in single cell data*.
Bioinformatics 38(20):4720–4726.
</div>

::: {.notes}
Step 3 is the part worth a sentence: with three donors a group you cannot
estimate a variance per cell type, so propeller pools that information across
cell types. Same trick as limma in bulk RNA-seq, same reason - not enough
replicates to go it alone.

Step 1 is the whole point. Everything else is bookkeeping on nine numbers.

Worth saying: this is the Phipson lab's method, and the practicals are adapted
from their workshop. CHECK the author list before claiming the dataset authors
are on it.
:::

## Reading the output

::: {style="font-size: 0.7em;"}
```{r propeller}
pr <- propeller(clusters = s2$celltype, sample = s2$sample, group = s2$group)
knitr::kable(pr[, c("PropMean.fetal", "PropMean.young", "PropMean.adult", "FDR")], digits = 2)
```
:::

- **`PropMean.*`** — mean *donor* proportion in each group
- **`FDR`** — Benjamini–Hochberg, across the `r nrow(pr)` cell types tested

::: {.fragment}
`r sum(pr$FDR < 0.05)` of `r nrow(pr)` cell types changed with development.
Neurons did not.
:::

## Compositional bias {background-color="#F5E9E9"}

Cardiomyocyte nuclei fell from **~74%** to **~24%**. Every other proportion
*has* to rise, whether or not anything happened to those cells.

<br>

::: {.fragment}
**propeller does not control for this.** It stabilises the variance and fits one
linear model per cell type. 
:::

::: {.fragment}
Drop cardiomyocytes, renormalise among the remaining five, rerun:

::: {style="font-size: 0.7em;"}
```{r comp-bias}
k <- s2$celltype != "Cardiomyocytes"
pr_nc <- propeller(clusters = s2$celltype[k], sample = s2$sample[k],
                   group = s2$group[k])
ct <- setdiff(rownames(pr), "Cardiomyocytes")
knitr::kable(data.frame(`FDR, all 6 types`  = pr[ct, "FDR"],
                        `FDR, without CM`   = pr_nc[ct, "FDR"],
                        row.names = ct, check.names = FALSE), digits = 3)
```
:::

:::

::: {.fragment}
Proportions only ever tell you about **relative** abundance. 
:::



# Part 3 Differential expression

## The question

Within cardiomyocytes: which genes differ between adult and foetal hearts?

<br>

Most commonly used: `Seurat::FindMarkers()`. On this data it returns
**~2,000 significant genes**.

::: {.fragment}
That number is inflated.
:::

## The problem of pseudoreplication

`FindMarkers` treats **every nucleus as an independent replicate**.

They are not. Nuclei from one donor share:

- that donor's genome
- their age, sex, medication, disease history
- one tissue dissociation, one library prep, one sequencing run

::: {.fragment}
<br>

**Effective sample size is 3 donors per group, not thousandsd of nuclei.**
:::



## The agent knew this 

One of the 'caveats' reported:

> "Wilcoxon on individual cells treats cells as replicates and **inflates
> significance** — for anything you'd report, redo this as donor-level
> pseudobulk."

::: {.fragment}
Its headline, same report: **"Adult vs. foetal cardiomyocytes: 5,967
significant genes"**. But it didn't do anything.
:::



## What can we do: pseudobulking

Sum counts within each **donor × cell type**. One profile per donor.

<br>

Then use methods built for a handful of replicates — which is what bulk RNA-seq
methods are. `edgeR`, `limma-voom`, `DESeq2`.

::: {.fragment}
<br>

Nine columns instead of seven thousand. **That was always the real sample size.**

We will see results in practical.
:::

<div style="font-size: 0.5em; color: #5f6b7a; margin-top: 0.8em;">
Squair et al. (2021), *Confronting false discoveries in single-cell differential
expression*. Nature Communications 12:5692.
</div>


# Part 4 Where the field is going

## Beyond discrete labels

**Trajectory / pseudotime** — order cells along a continuous process rather than
binning them. Slingshot, Monocle3, DPT, RNA velocity; our own **PhiSpace**.

**Cell–cell interaction** — infer signalling between cell types from
ligand–receptor expression. CellPhoneDB, CellChat.

**Cell-specific networks** — a co-expression network *per cell* rather than per
cluster. Our own **NeighbourNet**.



## Spatial transcriptomics

scRNA-seq tells you **which cells are doing what**.

Spatial tells you **where they are doing it**.

- Sequencing-based: Visium, Visium HD, Slide-seq
- Imaging-based: Xenium, CosMx, MERFISH

::: {.fragment}
Tumour microenvironment, tissue architecture, niches — questions dissociation
destroys.
:::


## Foundation models

scGPT, Geneformer, scFoundation, UCE — transformers pre-trained on tens of
millions of cells.

Promise: transfer to new tissues, few-shot annotation, in-silico perturbation.

::: {.fragment}
Open question: **do they beat well-tuned simple baselines?** For several tasks
the honest answer is still "not clearly".
:::

::: {.fragment}
This might be a training data issue.

::: {style="font-size: 0.55em; line-height: 1.35;"}
**Beating simple baselines:** Kedzierska, Crawford, Amini *et al.* (2025),
*Zero-shot evaluation reveals limitations of single-cell foundation models*.
Genome Biology 26:101.

**Batch effects:** Wang, Zhang & Zhang (2025), *Batch effects remain a
fundamental barrier to universal embeddings in single-cell foundation models*.
bioRxiv 2025.12.19.695371. *(preprint)*

**Training data scale:** DenAdel, Hughes, Thoutam *et al.* (2026),
*Evaluating the role of pretraining dataset size and diversity on single-cell
foundation model performance*. Nature Methods 23:1447–1457.
:::


:::




