---
title: "Practical 1 — From counts to clusters"
subtitle: "BINF90004 Bioinformatics Case Studies · scRNA-seq"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Before you start


```{r libraries}
library(Seurat)
library(harmony)
library(clustree)
library(ggplot2)
library(dplyr)
library(patchwork)

# Not optional. SCTransform hands large objects to parallel workers, and the
# default 500 MB ceiling is too low - without this you get a
# "total size of globals exceeds maximum allowed size" error partway through
# section 3, which looks alarming and is not your fault.
options(future.globals.maxSize = 8 * 1024^3)

# Resolve the data folder whether you knit from practicals/ or run
# interactively with the repo root as your working directory. The third
# branch is the instructor's private authoring folder.
DATA <- if (dir.exists("data")) "data" else
        if (dir.exists("../data")) "../data" else "../dist"
```

```{r palettes}
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")
sample_cols <- c(
    f1 = "#9ECAE1", f2 = "#6BAED6", f3 = "#3182BD",
    y1 = "#FDD0A2", y2 = "#FDAE6B", y3 = "#E6550D",
    a1 = "#C7E9C0", a2 = "#A1D99B", a3 = "#31A354"
)
```

---

# 1. The data

Human heart, single-**nucleus** RNA-seq, from
[Sim et al. (2021), *Circulation*](https://doi.org/10.1161/CIRCULATIONAHA.120.051921) —
a study of how the human heart matures. Nine donors across three developmental
stages:

| Group  | Donors | Age |
|--------|--------|-----|
| foetal | f1, f2, f3 | 19–20 weeks |
| young  | y1, y2, y3 | 4–14 years |
| adult  | a1, a2, a3 | 35–42 years |

Note **snRNA-seq**, not scRNA-seq: nuclei were isolated rather than whole cells.
Keep that in mind — it will matter within about ten minutes.

```{r load}
seu <- readRDS(file.path(DATA, "P1_start.rds"))
seu
```

Some work has been done for you, to save time — none of it involving a choice:
cells with <200 genes and genes in <3 cells were dropped, gene annotation was
attached, QC metrics were computed, uninformative gene classes (ribosomal,
unannotated, sex-chromosome) were removed, and the object was downsampled to
8,000 nuclei stratified by donor.

That last step preserves the donors' relative sizes, which are strikingly
uneven — the adult samples are genuinely much smaller than the foetal ones. That
imbalance is real, and it will matter twice today.

**No cell filtering has been done.** That is your first decision.

```{r metadata}
head(seu@meta.data)
table(seu$sample, seu$group)
```

Two metadata columns matter now:

- `nCount_RNA` — total UMI counts per nucleus (library size)
- `nFeature_RNA` — number of genes detected per nucleus

and two we computed:

- `percent.mt` — % of counts from mitochondrial genes
- `percent.ribo` — % of counts from ribosomal protein genes

---

# 2. Quality control

## 2.1 Look before you filter

The single most common QC mistake is applying a threshold you read somewhere
without first looking at your own data. So: look first.

```{r qc-violin, fig.width=11, fig.height=8}
VlnPlot(seu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
        group.by = "sample", pt.size = 0, ncol = 2) &
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Both panels below put library size on the x-axis. Before you look, know what each
one *should* show — a departure only means something if you had an expectation.

**Library size vs genes detected** should be a single tight cloud that rises and
then flattens, with all nine donors on the same curve. The two axes measure the
same underlying quantity — how much RNA was captured and sequenced — so they are
strongly correlated. The flattening is expected: detection is all-or-nothing per
gene, so extra depth increasingly re-sequences genes you have already seen. What
is informative is anything off that curve. Nuclei **below** it spread many counts
over few genes: a low-complexity library, either ambient RNA or a handful of
genes eating the depth. Nuclei far **up and to the right** of everything else are
candidate doublets, two nuclei captured in one droplet. The **bottom-left** cloud
is empty droplets and debris, and is what `min_counts` and `min_genes` remove. A
whole donor sitting on its own curve is not biology; it is a difference in
sequencing depth or prep.

**Library size vs % mitochondrial** is conventionally read as a cell-health plot,
and the reasoning is worth stating because you are about to test it. When a cell
dies, its membrane becomes permeable and cytoplasmic mRNA leaks out, while
mitochondria — which have their own double membrane — are retained. Damaged cells
therefore end up with *few counts* and a *high mitochondrial fraction*, giving a
characteristic L-shape: a dense band along the bottom, plus a tail climbing the
left-hand edge. That tail is what a `percent.mt` threshold exists to remove.

Keep that expected shape in mind when you look at the right-hand panel.

```{r qc-scatter, fig.width=11, fig.height=4.5}
p1 <- FeatureScatter(seu, "nCount_RNA", "nFeature_RNA", group.by = "sample",
                     cols = sample_cols) + ggtitle("Library size vs genes detected")
p2 <- FeatureScatter(seu, "nCount_RNA", "percent.mt", group.by = "sample",
                     cols = sample_cols) + ggtitle("Library size vs % mitochondrial")
p1 + p2
```

Before reading on, answer two questions from the plots:

1. Do all nine donors look alike?
2. Where would you put a mitochondrial cut-off?

## 2.2 What `percent.mt` actually looks like here

```{r pmt-summary}
summary(seu$percent.mt)

# Per-donor: what fraction of nuclei would each candidate threshold remove?
sapply(c(1, 5, 10, 20), function(t) {
    round(100 * tapply(seu$percent.mt, seu$sample, function(z) mean(z >= t)), 1)
}) |> `colnames<-`(paste0("< ", c(1, 5, 10, 20)))
```

The median is around **0.25%**. That is very low — and it is low *because these
are nuclei*. Mitochondria live in the cytoplasm, and the nuclei prep washes the
cytoplasm away. A threshold carried over from a whole-cell dataset is answering a
question this assay does not ask.

So the L-shape is largely absent, and with it the interpretation that justified
the threshold. What `percent.mt` still tracks in a nuclei prep is how completely
the cytoplasm was washed away — a property of the *preparation*, not of the
health of any individual nucleus. That has a consequence which matters in
Practical 2: a metric driven by prep does not vary evenly across the things you
want to compare. It varies by cell type — cardiomyocytes carry the **lowest**
mitochondrial fraction of any type in this dataset, so cutting on `percent.mt`
preferentially removes everything that is *not* a cardiomyocyte. Note that the
intuitive argument here — heart tissue is mitochondria-rich, so expect a high
`percent.mt` — gets this backwards; you can check the claim yourself once you
have annotated clusters in Practical 2. And it varies by donor.

Now look across the rows of that table rather than down the columns.

> ### <span style="color:#C0392B">DECISION 1 — where do you cut? *(10 min)*</span>
>
> Choose values for `min_genes`, `min_counts`, `max_counts` and `max_mt` below.
>
> Then answer, in writing, and be ready to defend it:
>
> 1. **How many nuclei did you lose, and from which donors?** Is the loss even?
> 2. A widely-copied tutorial default is `percent.mt < 5`. Look at that column in
>    the table above. It removes about **6%** of nuclei overall — but **18% of
>    donor `a3`** and **1.5% of donor `f1`**, a twelve-fold difference. Is that
>    acceptable? What would it do to a later comparison of cell type composition
>    between developmental stages — which is exactly what you will run in
>    Practical 2?
> 3. Now look at the `< 1` column, where donor `y3` loses 61% of its nuclei and
>    `f1` loses 9%. At what point does a quality threshold stop measuring quality
>    and start measuring which donor a nucleus came from?
> 4. Would you use one global threshold, or a per-donor one? What is the cost of
>    each?
>


```{r decision-1}
# ---- YOUR VALUES ----
min_genes  <- 500
min_counts <- 2500
max_counts <- 40000
max_mt     <- 20
# ---------------------

n_before <- ncol(seu)
seu_filt <- subset(seu, subset = nFeature_RNA > min_genes &
                                 nCount_RNA  > min_counts &
                                 nCount_RNA  < max_counts &
                                 percent.mt  < max_mt)

cat("nuclei:", n_before, "->", ncol(seu_filt),
    sprintf("(%.1f%% removed)\n", 100 * (1 - ncol(seu_filt) / n_before)))

# Where did the losses fall? Index by name, so this still works if a donor
# loses every nucleus at an aggressive threshold.
before <- table(seu$sample)
after  <- table(factor(seu_filt$sample, levels = names(before)))
data.frame(
    before   = as.integer(before),
    after    = as.integer(after),
    pct_lost = round(100 * (1 - as.integer(after) / as.integer(before)), 1),
    row.names = names(before)
)
```

> ### <span style="color:#2980B9">🤖 What the agent chose here</span>
>
> Given this dataset and **no instructions at all**, Claude Science picked
> **≥500 genes, ≥1000 UMI, `percent.mt < 10`** — and justified the last one like
> this:
>
> > *"mitochondrial fractions are uniformly low (donor medians 0.10–1.78%) **and**
> > so are ribosomal fractions (0.4–1.3%). That combination points to a
> > single-nucleus prep rather than whole cells."*
>
> Nobody told it this was snRNA-seq. It worked that out from the data. It also
> flagged that donor `y3` has the shallowest library and that `a3` ends up with
> only ~200 cells — the same donor problem you just found.
>
> **Your question:** can you justify your threshold as well as that? Not "is your
> number the same" — is your *reason* as good?

> **<span style="color:#C0392B">HARD SYNC POINT.</span>** Whatever you chose, run the
> chunk below now — including if your own filtering never ran. From here everyone
> works on an identical object, so the rest of the session lines up. Your own
> thresholds stay in your notes; we come back to them in the wrap-up.

```{r sync-1}
# Reference thresholds. Re-read from the original file so this works no matter
# what state your object is in.
seu <- readRDS(file.path(DATA, "P1_start.rds"))
seu <- subset(seu, subset = nFeature_RNA > 500 & nCount_RNA > 2500 &
                            nCount_RNA < 40000 & percent.mt < 20)
seu
```

---

# 3. Normalisation

Nuclei differ enormously in total counts, for reasons that have nothing to do
with biology — capture efficiency, RNA content, sequencing depth. Normalisation
removes that technical variation.

**SCTransform** models each gene's counts with a regularised negative binomial
and returns Pearson residuals. It handles the mean–variance relationship (highly
expressed genes vary more, just from sampling) better than a simple log
transform, and picks variable features as a side effect.

```{r sctransform}
seu <- SCTransform(seu, vst.flavor = "v2", seed.use = 42, verbose = FALSE)
```

```{r hvg}
head(VariableFeatures(seu), 20)
```

---

# 4. Dimension reduction

```{r pca}
seu <- RunPCA(seu, seed.use = 42, verbose = FALSE)
```

```{r elbow, fig.width=7, fig.height=4.5}
ElbowPlot(seu, ndims = 40)
```

> ### <span style="color:#C0392B">DECISION 2 — how many PCs? *(8 min)*</span>
>
> Pick `n_dims` from the elbow plot.
>
> 1. Where is the elbow? Do you and your neighbour agree to within 5?
> 2. What happens downstream if you take too **few**? Too **many**?
> 3. The elbow plot is a heuristic with no theoretical justification. What would
>    a more principled choice look like, and why do you think almost nobody
>    bothers?

```{r decision-2}
n_dims <- 20   # <- YOUR VALUE
```

---

# 5. Integration

Cells frequently cluster by *donor* rather than by *cell type*, because each
sample carries its own technical signature. Look for that first:

```{r pca-batch, fig.width=7, fig.height=5}
DimPlot(seu, reduction = "pca", group.by = "sample", cols = sample_cols,
        shuffle = TRUE) + ggtitle("PCA — before integration")
```

**Harmony** iteratively projects cells into a shared embedding, clustering them
softly and shifting each cluster to remove the donor-associated component while
leaving the rest alone.

```{r harmony}
set.seed(42)
seu <- RunHarmony(seu, group.by.vars = "sample", reduction.use = "pca",
                  dims.use = 1:n_dims, verbose = FALSE)
```

```{r harmony-compare, fig.width=11, fig.height=5}
p1 <- DimPlot(seu, reduction = "pca", group.by = "sample", cols = sample_cols,
              shuffle = TRUE) + ggtitle("Before") + NoLegend()
p2 <- DimPlot(seu, reduction = "harmony", group.by = "sample", cols = sample_cols,
              shuffle = TRUE) + ggtitle("After (Harmony)")
p1 + p2
```

**A warning worth more than the method.** "The donors mix better" is also what
over-correction looks like — and over-correction destroys real biology. Here the
donors differ by developmental stage, so *some* separation between them is
exactly the signal we are trying to study. Integrating by `sample` rather than by
`group` is a deliberate choice to remove donor effects while preserving stage
effects.

So: can you check whether it worked?

## 5.1 A test for over-correction

> **Where this came from.** This check was not in the original version of this
> practical. It came from watching an AI agent analyse this exact dataset with no
> instructions — it invented this test, and it is better than the warning we used
> to give you here. Materials improve from wherever the good idea comes.

The argument has two halves, and you need both:

1. **Did integration remove the donor effect?** Count clusters dominated by a
   single donor, before and after. Should fall.
2. **Did it also remove the biology?** Count clusters restricted to a single
   developmental *stage*. Should **not** fall — because that is the signal.

If clusters stop being donor-specific but stay stage-specific, and each of those
stage-specific clusters now draws on all three donors of that stage, the
remaining structure is developmental rather than technical.

```{r overcorrection-check}
# Self-contained: clusters both embeddings at a fixed resolution, so this does
# not depend on the resolution you pick in Decision 3. About 3 seconds.
check_res <- 0.4

dominated <- function(clusters, by, thr = 0.9) {
    tb <- table(clusters, by)
    frac <- apply(tb, 1, max) / rowSums(tb)
    c(clusters   = nrow(tb),
      dominated  = sum(frac >= thr),
      median_grp = median(apply(tb, 1, function(r) sum(r > 0))))
}

tmp <- FindNeighbors(seu, reduction = "pca", dims = 1:n_dims,
                     graph.name = c("p_nn", "p_snn"), verbose = FALSE)
tmp <- FindClusters(tmp, graph.name = "p_snn", resolution = check_res, verbose = FALSE)
before <- Idents(tmp)

tmp <- FindNeighbors(seu, reduction = "harmony", dims = 1:n_dims,
                     graph.name = c("h_nn", "h_snn"), verbose = FALSE)
tmp <- FindClusters(tmp, graph.name = "h_snn", resolution = check_res, verbose = FALSE)
after <- Idents(tmp)

rbind(`before integration` = dominated(before, seu$sample),
      `after Harmony`      = dominated(after,  seu$sample))
```

Donor-dominated clusters should collapse towards zero. Now the other half — did
the biology survive?

```{r stage-restriction}
tb <- table(after, seu$group)
cat("clusters >=90% one developmental stage, after integration: ",
    sum(apply(tb, 1, max) / rowSums(tb) >= 0.9), " of ", nrow(tb), "\n", sep = "")
rm(tmp)
```

Donor structure gone, stage structure intact. Unlike "the UMAP looks better",
that is a pair of numbers you can put in a methods section.

```{r umap}
seu <- RunUMAP(seu, reduction = "harmony", dims = 1:n_dims,
               seed.use = 42, verbose = FALSE)
```

```{r umap-plots, fig.width=11, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", group.by = "sample", cols = sample_cols,
              shuffle = TRUE) + ggtitle("By donor")
p2 <- DimPlot(seu, reduction = "umap", group.by = "group", cols = group_cols,
              shuffle = TRUE) + ggtitle("By developmental stage")
p1 + p2
```

Remember from the lecture: UMAP is for looking, not for measuring. Distances
between clusters mean little, cluster areas mean nothing at all, and a cluster
that looks isolated may not be.

---

# 6. Clustering

`FindNeighbors()` builds a shared-nearest-neighbour graph; `FindClusters()` runs
Louvain community detection on it. The `resolution` parameter controls
granularity, and it has no correct value — it depends on the question.

```{r neighbours}
seu <- FindNeighbors(seu, reduction = "harmony", dims = 1:n_dims, verbose = FALSE)
```

```{r sweep}
resolutions <- c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8)
set.seed(42)
for (r in resolutions) seu <- FindClusters(seu, resolution = r, verbose = FALSE)

sapply(resolutions, function(r)
    length(unique(seu@meta.data[[paste0("SCT_snn_res.", r)]]))) |>
    setNames(paste0("res ", resolutions))
```

```{r clustree, fig.width=10, fig.height=9}
clustree(seu, prefix = "SCT_snn_res.")
```

Reading a clustree: resolution increases downward. **Thick straight edges** mean
a cluster is stable — the same cells stay together as resolution rises.
**Thin, criss-crossing edges** mean cells are being shuffled between clusters,
which usually means you are splitting a continuous population that has no natural
division.

> ### <span style="color:#C0392B">DECISION 3 — how many clusters are real? *(11 min)*</span>
>
> Choose `selected_res`.
>
> 1. Find a cluster in the clustree that splits cleanly, and one that shuffles.
>    What is the difference telling you?
> 2. Look at the cluster counts above. Even at the **lowest** resolution you
>    tried, you get about **12 clusters**. The published analysis of this tissue
>    describes **8 cell types**. Resolution 0.1 is not "too coarse" — it already
>    over-splits. What does that tell you about treating a cluster as a cell type?
> 3. Resolutions 0.4 and 0.6 give the *same* number of clusters. Are they the
>    same clustering? How would you check?
> 4. **The hard one.** What evidence, other than the clustree, could tell you
>    that a cluster is real? Note that you have not looked at a single gene yet.
> 5. If your answer to (4) is "check the marker genes" — you will do exactly that
>    in Practical 2, and `FindAllMarkers` will return markers for *whatever*
>    clusters you hand it, including ones you invented by turning a dial. Does
>    that worry you?

```{r decision-3}
selected_res <- 0.4   # <- YOUR VALUE

Idents(seu) <- paste0("SCT_snn_res.", selected_res)
seu$seurat_clusters <- Idents(seu)
cat("clusters:", length(unique(seu$seurat_clusters)), "\n")
```

```{r umap-clusters, fig.width=11, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", label = TRUE, repel = TRUE) +
    NoLegend() + ggtitle(paste("Clusters (res =", selected_res, ")"))
p2 <- DimPlot(seu, reduction = "umap", group.by = "group", cols = group_cols,
              shuffle = TRUE) + ggtitle("Developmental stage")
p1 + p2
```

```{r cluster-composition, fig.width=9, fig.height=5}
seu@meta.data |>
    count(seurat_clusters, group) |>
    group_by(seurat_clusters) |>
    mutate(prop = n / sum(n)) |>
    ggplot(aes(seurat_clusters, prop, fill = group)) +
    geom_col() +
    scale_fill_manual(values = group_cols) +
    labs(x = "cluster", y = "proportion of cluster", fill = NULL) +
    theme_minimal()
```

> ### <span style="color:#2980B9">🤖 What the agent chose here</span>
>
> **Leiden at resolution 0.6**, against your Louvain at
> `r if (exists("selected_res")) selected_res else 0.4`. Different algorithm,
> different resolution, different cluster count — on identical data.
>
> Neither is wrong. But notice that nothing in either analysis reports *"this
> number was a choice, and here is what it changes."* You only know because you
> are holding both.

Several clusters are almost entirely foetal. That is real developmental biology,
not a batch effect — foetal heart runs transcriptional programmes that the adult
heart has switched off. It is also a warning: a clustering algorithm that splits
strongly on developmental stage will happily split on any other strong covariate,
including a technical one.

---

# 7. Save

```{r save, eval=FALSE}
dir.create("results", showWarnings = FALSE)
saveRDS(seu, "results/my_02_integrated_clustered.rds")
```

Practical 2 starts from the shared checkpoint, not from your file, so the two
sessions stay in step:

```{r p2-entry, eval=FALSE}
seu <- readRDS(file.path(DATA, "checkpoints/02_integrated_clustered.rds"))
```

---

# What you decided

Write these down — we come back to them at the end of the day, and they are the
shape of what your report needs to argue.

| | Your value | Why |
|---|---|---|
| QC thresholds | | |
| Number of PCs | | |
| Clustering resolution | | |

The pipeline ran identically whatever you chose. It would also have run
identically if you had chosen badly. **Nothing in the output tells you which.**

```{r session}
sessionInfo()
```
