---
title: "Practical 2 — From clusters to conclusions"
subtitle: "BINF90004 Bioinformatics Case Studies · scRNA-seq"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Before you start

**65 minutes, one hard sync point, two decisions.**

In Practical 1 you turned a count matrix into 19 clusters. Clusters are not an
answer — they are numbered blobs. Today you give them names, and then use those
names to make two biological claims: *which cell types change in abundance as the
heart matures*, and *which genes change within a cell type*.

The second of those is where most published single-cell analyses go wrong, and it
is the main thing to take away from today.

```{r libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(speckle)
library(edgeR)
library(limma)

options(future.globals.maxSize = 8 * 1024^3)

# Resolve the data folder whether you knit from practicals/ or run
# interactively with the repo root as your working directory.
DATA <- if (dir.exists("data")) "data" else "../data"
```

```{r palettes}
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")
```

```{r load}
seu <- readRDS(file.path(DATA, "checkpoints/02_integrated_clustered.rds"))
seu
table(seu$seurat_clusters)
```

---

# 1. Marker genes

A cluster is defined by the genes that are higher in it than in everything else.
`FindAllMarkers()` runs that comparison for every cluster.

**We are loading a cached result rather than computing it.** Seurat's default
Wilcoxon implementation takes several minutes on 19 clusters, and there is no
teaching value in watching a progress bar. The exact call that produced the
cache is in the chunk below — run it later if you want.

```{r markers}
markers <- readRDS(file.path(DATA, "markers_cache.rds"))

# How it was produced:
# seu <- PrepSCTFindMarkers(seu)
# markers <- FindAllMarkers(seu, only.pos = TRUE,
#                           min.pct = 0.25, logfc.threshold = 0.5)

nrow(markers)
head(markers)
```

```{r top-markers}
top5 <- markers |>
    group_by(cluster) |>
    slice_max(avg_log2FC, n = 5) |>
    summarise(top_genes = paste(gene, collapse = ", "))
top5
```

## 1.1 Canonical markers

Published markers for the major cardiac populations (Sim et al. 2021 and
standard references). Dot size is the fraction of nuclei expressing the gene;
colour is mean expression.

```{r canonical-panel}
panel <- list(
    Cardiomyocyte  = c("TNNT2", "ACTN2", "TNNI3", "RYR2", "MYH6", "TTN"),
    Fibroblast     = c("COL1A1", "COL3A1", "DCN", "PDGFRA"),
    Endothelial    = c("PECAM1", "CDH5", "VWF", "FLT1"),
    Immune         = c("PTPRC", "CD68", "CD163", "CD74"),
    Epicardial     = c("UPK3B", "WT1", "MSLN"),
    Neural         = c("PLP1", "S100B", "NRXN1"),
    `Smooth muscle`= c("MYH11", "TAGLN", "MYLK")
)
genes <- unlist(panel, use.names = FALSE)
genes <- genes[genes %in% rownames(seu)]   # drop anything filtered out earlier
setdiff(unlist(panel, use.names = FALSE), genes)  # what was missing, if any
```

```{r dotplot, fig.width=13, fig.height=6}
DotPlot(seu, features = genes, assay = "SCT") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(x = NULL, y = "cluster")
```

---

# 2. Naming the clusters

> ### <span style="color:#C0392B">DECISION 4 — what is your cluster? *(15 min)*</span>
>
> Your demonstrator will give each pair **one cluster number**.
>
> 1. Pull its top markers with the chunk below.
> 2. Look them up. [PanglaoDB](https://panglaodb.se/search.html) is a searchable
>    database of cell-type markers; the [Human Protein Atlas](https://www.proteinatlas.org/)
>    is good for single genes.
> 3. **Then go back and check your cluster in the DotPlot above.** Do not skip
>    this. `FindAllMarkers` returns what makes your cluster *different from the
>    other eighteen* — which is not the same as what makes it what it is. If
>    eleven clusters are the same cell type, the genes defining that cell type
>    cancel out and never appear in any of their marker lists.
> 4. **Commit to an answer before you look at the reveal.** Write down the cell
>    type *and* your confidence: certain / fairly sure / guessing.
> 5. Then answer: what would have changed your mind? If nothing would have, you
>    were not really testing a hypothesis.

```{r my-cluster}
my_cluster <- 5   # <- YOUR CLUSTER NUMBER

markers |>
    filter(cluster == my_cluster) |>
    slice_max(avg_log2FC, n = 15) |>
    select(gene, avg_log2FC, pct.1, pct.2, p_val_adj)
```

```{r my-cluster-umap, fig.width=11, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
p2 <- DimPlot(seu, reduction = "umap",
              cells.highlight = colnames(seu)[seu$seurat_clusters == my_cluster],
              cols.highlight = "#C0392B") +
    NoLegend() + ggtitle(paste("Cluster", my_cluster))
p1 + p2
```

## 2.1 The reveal

These are the labels from the published Sim et al. analysis.

> ### <span style="color:#C0392B">"Published" is not "correct"</span>
>
> These labels are one careful team's interpretation, produced by people who
> faced the same choices you just did. In at least two clusters here a
> marker-based reading is *more specific* than the published label, and in at
> least one the markers point somewhere the label does not.
>
> **If your answer disagrees, work out why before assuming you were wrong.**

```{r reveal}
key <- readRDS(file.path(DATA, "celltype_key.rds"))
tab <- table(cluster = seu$seurat_clusters, published = key[colnames(seu)])

# For your cluster: what did the original authors call these cells?
sort(tab[as.character(my_cluster), ], decreasing = TRUE) |> head(3)
```

```{r reveal-all}
data.frame(
    cluster   = rownames(tab),
    n         = as.integer(rowSums(tab)),
    published = colnames(tab)[apply(tab, 1, which.max)],
    agreement = paste0(round(100 * apply(tab, 1, max) / rowSums(tab)), "%")
)
```

**Look at that table as a whole before moving on.** Two things are wrong with it,
in opposite directions.

**Too fine.** You produced **19 clusters**, and **eleven of them are
cardiomyocytes**. That is not a failure — cardiomyocytes genuinely differ across
foetal, young and adult hearts, and the algorithm is finding real structure. But
"number of clusters" answered a question nobody asked. The resolution you chose
in Practical 1 decided how finely one cell type got subdivided, and **no marker
analysis would have warned you**, because each of those eleven clusters has
perfectly good markers.

**Too coarse.** The published analysis describes **8** cell types. Your 19
clusters recover **6**. Run this:

```{r rare-types}
key_tab <- sort(table(key[colnames(seu)]), decreasing = TRUE)
key_tab
```

**Smooth muscle cells** (62 nuclei) and **erythroid cells** (9 nuclei) are in
your data, and no cluster corresponds to either. Sixty of the 62 smooth muscle
nuclei are sitting inside cluster 3 — which is why cluster 3's agreement with the
published labels is the worst on the board:

```{r smooth-muscle}
sm <- names(key)[key == "Smooth muscle cells"]
table(seu$seurat_clusters[colnames(seu) %in% sm])
```

So the same analysis simultaneously split one cell type into eleven and merged
two others into invisibility. **Nothing in the output flagged either.** Rare
populations are the ones most likely to be lost, and — as you saw with the QC
thresholds in Practical 1 — also the ones most likely to be filtered away
earlier. Both failures point the same direction.

> ### <span style="color:#2980B9">🤖 What the agent found here</span>
>
> Claude Science scored **eleven** canonical categories against its clusters:
> cardiomyocyte, endothelial, fibroblast, pericyte/SMC, myeloid, lymphoid, mast,
> neural/glial, epicardial, erythroid, cycling.
>
> Note what is on that list that is **not** in your six: pericyte/smooth muscle,
> mast cells, erythroid — and it split immune into myeloid vs lymphoid, which is
> your clusters 6 and 16.
>
> It looked for the rare populations *by name*, because it started from a
> reference panel rather than from the clusters. You started from the clusters
> and asked what they were. **Those two directions fail differently:** the panel
> can find only what is on it; the cluster-first approach loses whatever failed
> to form a cluster.

> **<span style="color:#C0392B">HARD SYNC POINT.</span>** Run this now regardless of
> where you got to. It applies the reference labels so everyone's downstream
> numbers match.

```{r sync-2}
labels <- readRDS(file.path(DATA, "cluster_labels.rds"))
seu$celltype <- unname(labels[as.character(seu$seurat_clusters)])
table(seu$celltype)
```

```{r annotated-umap, fig.width=12, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", group.by = "celltype", label = TRUE,
              repel = TRUE) + NoLegend() + ggtitle("Cell type")
p2 <- DimPlot(seu, reduction = "umap", group.by = "group", cols = group_cols,
              shuffle = TRUE) + ggtitle("Developmental stage")
p1 + p2
```

---

# 3. Does composition change with development?

The first biological question: as the heart matures, do the *proportions* of cell
types change?

```{r composition, fig.width=10, fig.height=5}
comp <- seu@meta.data |>
    count(sample, group, celltype) |>
    group_by(sample) |>
    mutate(prop = n / sum(n)) |>
    ungroup()

ggplot(comp, aes(sample, prop, fill = celltype)) +
    geom_col() +
    facet_grid(~ group, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "proportion of nuclei", fill = NULL) +
    theme_minimal()
```

```{r composition-box, fig.width=11, fig.height=5}
ggplot(comp, aes(group, prop, fill = group)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, size = 1.2) +
    facet_wrap(~ celltype, scales = "free_y", nrow = 2) +
    scale_fill_manual(values = group_cols) +
    labs(x = NULL, y = "proportion") +
    theme_minimal() + theme(legend.position = "none")
```

It is tempting to test this with a chi-square test on the cell counts. **Do not.**
That treats every nucleus as an independent observation, so with ~7,000 nuclei
almost everything comes out significant. Your independent units are the **nine
donors**, not the nuclei.

`propeller` does it properly: it computes each donor's proportions, transforms
them (proportions are bounded and heteroscedastic), and tests across donors.

```{r propeller}
prop_res <- propeller(clusters = seu$celltype, sample = seu$sample,
                      group = seu$group)
prop_res
```

This is a strong result even at n = 3 per group. Cardiomyocytes fall from about
**74% of nuclei in foetal hearts to 24% in adult**; immune cells and fibroblasts
rise correspondingly. Neurons do not change.

Two cautions before you believe it:

1. **These are nuclear proportions, not cell proportions.** Adult cardiomyocytes
   are frequently binucleated and polyploid, and non-myocytes keep dividing. Some
   of this shift is real change in tissue composition, and some is a change in
   how many nuclei a cardiomyocyte contributes. A proportion is only as
   interpretable as its denominator.
2. **Compositions are constrained to sum to 1.** If cardiomyocytes fall, then
   *everything else must rise* whether or not anything happened to it. You cannot
   read the fibroblast result independently of the cardiomyocyte result.

Recall too from Practical 1 that donor `a3` contributes far fewer nuclei than the
others, so its proportions are the noisiest of the nine — and it is one of only
three adults.

> ### <span style="color:#2980B9">🤖 What the agent did here</span>
>
> It reported cardiomyocytes falling from **64.4%** of cells in foetal hearts to
> **23.3%** in adult — essentially the same numbers you just produced — and
> described the change as developmental biology.
>
> **It ran no statistical test of composition at all.** No propeller, no scCODA,
> no chi-square. Nothing.
>
> And here is the part worth sitting with: it wrote a careful, honest caveats
> section listing four limitations of its analysis. **This was not one of them.**
>
> You cannot catch a missing analysis by reading the limitations. You catch it by
> knowing what should have been there.

---

# 4. Differential expression — the important part

Second question: within cardiomyocytes, which genes change between foetal and
adult hearts?

## 4.1 The tempting approach

Every scRNA-seq tutorial reaches for `FindMarkers()`. Let us run it — but on a
**random 100 nuclei per group**, so it finishes in about ten seconds.

```{r cm-subset}
cm <- subset(seu, subset = celltype == "Cardiomyocytes" &
                           group %in% c("adult", "fetal"))
table(cm$group, cm$sample)
```

```{r naive-de-small}
set.seed(42)
cells <- unlist(lapply(c("adult", "fetal"), function(g)
    sample(colnames(cm)[cm$group == g], 100)))
small <- subset(cm, cells = cells)

Idents(small) <- "group"
naive_small <- FindMarkers(small, ident.1 = "adult", ident.2 = "fetal",
                           assay = "SCT", verbose = FALSE)
sum(naive_small$p_val_adj < 0.05)
```

Several hundred significant genes, from 200 nuclei.

Now: **before running the next chunk, predict.** If we repeat this with all
~3,200 cardiomyocytes instead of 200, what happens to that number? Write down a
guess.

## 4.2 A control that should find nothing

Rather than scaling up, let us do something more useful: run **exactly the same
test on a comparison where the right answer is known to be "nothing"**.

Take two foetal donors. Same developmental stage, same tissue, same protocol.
There is no developmental difference between `f1` and `f2` — they are two
instances of the same biological condition. A test asking "do these differ?"
should come back nearly empty.

These are pre-computed (about four minutes to run), all at the **same 300 cells
per side**, so the numbers are directly comparable.

```{r de-demo}
demo <- readRDS(file.path(DATA, "de_demo.rds"))
demo$wilcoxon
```

```{r demo-plot, fig.width=7.5, fig.height=4}
ggplot(demo$wilcoxon, aes(reorder(comparison, n_sig), n_sig, fill = kind)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(`real contrast` = "#2980B9",
                                 `same stage (foetal)` = "#C0392B")) +
    labs(x = NULL, y = "genes called significant (padj < 0.05)", fill = NULL,
         title = "Per-nucleus Wilcoxon, 300 cells per side",
         subtitle = "the red bars are comparisons with no biology in them") +
    theme_minimal()
```

Two donors at the same developmental stage produce **hundreds of "significantly
differentially expressed" genes**. There is no developmental biology in that
comparison. Every one of those genes is donor variation — genetics, age, tissue
handling, dissociation, library prep — being reported to you as a finding.

And that noise floor is a substantial fraction of the "real" adult-vs-foetal
result from the same test at the same cell count.

**So when the per-nucleus test tells you 2,000 genes separate adult from foetal
cardiomyocytes, it cannot distinguish "adult differs from foetal" from "these
three donors differ from those three donors."** It has no way to tell them apart,
because it never modelled donor at all.

Now try the same control with pseudobulk and you hit something instructive:

```{r pb-null, eval=FALSE}
# f1 vs f2, pseudobulk: one sample per group.
# limma cannot fit this - there is no residual variance to estimate.
```

It refuses. **That is the correct answer.** You cannot test a difference between
two groups with one sample in each. The per-nucleus test does not refuse — it
manufactures thousands of degrees of freedom out of one donor's nuclei and hands
back a gene list.

## 4.3 Why the naive test is wrong

That test treated every nucleus as an independent replicate. It is not. Nuclei
from the same donor share that donor's genetics, age, medication, tissue handling
and library prep — they are **subsamples of one biological observation**, not
independent ones.

This is **pseudoreplication**, and it is the single most common statistical error
in the single-cell literature (Squair et al., *Nat Commun* 2021). The effective
sample size is not thousands of nuclei. It is **three donors per group.**

Inflating n inflates significance. A p-value computed on 3,800 nuclei is
answering "are these two piles of nuclei different?", which they always are.
The question you actually asked was "do adult hearts differ from foetal hearts?"

## 4.3 The defensible approach: pseudobulk

Sum counts within each donor × cell type, giving one profile per donor — then use
methods built for a handful of replicates, which is exactly what bulk RNA-seq
methods are.

```{r pseudobulk}
seu$Celltype <- seu$celltype
seu$Sample   <- seu$sample
DefaultAssay(seu) <- "RNA"

pb <- Seurat2PB(seu, sample = "Sample", cluster = "Celltype")
dim(pb)

# Cardiomyocytes only
pb_cm <- pb[, pb$samples$cluster == "Cardiomyocytes"]
pb_cm$samples$group <- sub("[0-9]+$", "", pb_cm$samples$sample)
pb_cm$samples$group <- c(a = "adult", f = "fetal", y = "young")[pb_cm$samples$group]
table(pb_cm$samples$group)
```

Nine columns. That is the real sample size, and it was always the real sample
size.

```{r pb-de}
keep <- filterByExpr(pb_cm, group = pb_cm$samples$group)
pb_cm <- pb_cm[keep, , keep.lib.sizes = FALSE]
pb_cm <- normLibSizes(pb_cm)

design <- model.matrix(~ 0 + factor(pb_cm$samples$group))
colnames(design) <- levels(factor(pb_cm$samples$group))

v   <- voom(pb_cm, design, plot = TRUE)
fit <- lmFit(v, design)
fit <- contrasts.fit(fit, makeContrasts(adult - fetal, levels = design))
fit <- eBayes(fit)

summary(decideTests(fit))
```

```{r pb-top}
topTable(fit, n = 15)
```

```{r volcano, fig.width=7, fig.height=5.5}
tt <- topTable(fit, n = Inf)
tt$sig <- tt$adj.P.Val < 0.05 & abs(tt$logFC) > 1
ggplot(tt, aes(logFC, -log10(adj.P.Val), colour = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#C0392B")) +
    labs(title = "Cardiomyocytes: adult vs foetal (pseudobulk)",
         x = "log2 fold change", y = "-log10 adjusted p") +
    theme_minimal() + theme(legend.position = "none")
```

> ### <span style="color:#2980B9">🤖 What the agent did here</span>
>
> Same contrast, same data. It ran **per-cell Wilcoxon** and reported
> **5,967 significant genes** as its headline result, with a volcano plot.
>
> Then, in its caveats:
>
> > *"Wilcoxon on individual cells treats cells as replicates and **inflates
> > significance** — for anything you'd report, redo this as donor-level
> > pseudobulk with DESeq2 at n=3 per stage."*
>
> It is completely right. It diagnosed the problem you have just spent twenty
> minutes on, unprompted, in writing.
>
> **And it did not do it.** Pseudobulk appears in its output only as a heatmap —
> a picture, not a test. The number in the headline is the one it told you not to
> use.
>
> **The failure is not ignorance. It is that the caveat and the number do not
> carry equal weight** — and a reader who skims takes away 5,967.

> ### <span style="color:#C0392B">DECISION 5 — which number do you report? *(10 min)*</span>
>
> You now have two answers to the same biological question.
>
> 1. Write both counts down. What is the ratio?
> 2. Both analyses are "correct" in that the code ran and the maths is right.
>    What makes one of them *wrong*?
> 3. The per-nucleus test is faster, gives more hits, and appears in hundreds of
>    published papers. Given that, why should you use the other one?
> 4. Suppose a reviewer asks you to justify your choice. Write the sentence you
>    would put in the response. That sentence is the deliverable — not the gene
>    list.

---

# 5. What this analysis cannot tell you

Worth being explicit, because your report will need a section like this:

- **n = 3 per group.** Small. `propeller` finding little is a statement about
  power as much as about biology.
- **Sex is confounded with stage** (foetal 2M/1F, young 2M/1F, adult 1M/2F).
  Sex-chromosome genes were removed, which helps but does not fix an unbalanced
  design.
- **Donors differ in depth and quality.** You saw this in Practical 1: `a3` loses
  far more nuclei to any QC threshold than `f1`. Those two facts are connected.
- **Cell type labels are an interpretation**, not a measurement. You chose a
  clustering resolution; that choice produced eleven cardiomyocyte clusters that
  were then collapsed back into one label. A different resolution gives different
  boundaries and therefore different pseudobulk profiles.
- **Pseudobulk discards within-cell-type heterogeneity** by construction. If the
  interesting biology is a shift in the *distribution* of cell states rather than
  in the mean, this analysis will not see it.

---

# The five decisions

| | Your value | What a default would have done |
|---|---|---|
| 1. QC thresholds | | `percent.mt < 5` |
| 2. Number of PCs | | 10 |
| 3. Clustering resolution | | 0.8 |
| 4. Cluster identity | | — |
| 5. DE method | | per-nucleus Wilcoxon |

Every one of these was a choice. None of them was made for you by the data, and
none of them would have produced an error message if you had chosen badly.

```{r session}
sessionInfo()
```
