---
title: "Practical 2 — From clusters to conclusions"
subtitle: "BINF90004 Bioinformatics Case Studies · scRNA-seq"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

In Practical 1 we identified 19 clusters. We still need to make sense of these clusters -- by giving them cell type labels. We then use the labelled, or annotated, clusters to make two biological claims: *which cell types change in abundance as the heart matures*, and *which genes change within a cell type*.

```{r libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(speckle)
library(edgeR)
library(limma)

options(future.globals.maxSize = 8 * 1024^3)

# Resolve the data folder whether you knit from practicals/ or run
# interactively with the repo root as your working directory. The third
# branch is the instructor's private authoring folder.
DATA <- if (dir.exists("data")) "data" else
        if (dir.exists("../data")) "../data" else "../dist"
```

```{r palettes}
group_cols <- c(fetal = "#4E79A7", young = "#F28E2B", adult = "#59A14F")
```

```{r load}
seu <- readRDS(file.path(DATA, "checkpoints/02_integrated_clustered.rds"))
seu
table(seu$seurat_clusters)
```

---

# 1. Marker genes

A cluster is defined by the genes that are higher in it than in everything else.
`FindAllMarkers()` runs that comparison for every cluster.

**We are loading a cached result rather than computing it.** Seurat's default
Wilcoxon implementation takes several minutes on 19 clusters, and there is no
teaching value in watching a progress bar. The exact call that produced the
cache is in the chunk below — run it later if you want.

```{r markers}
markers <- readRDS(file.path(DATA, "markers_cache.rds"))

# How it was produced:
# seu <- PrepSCTFindMarkers(seu)
# markers <- FindAllMarkers(seu, only.pos = TRUE,
#                           min.pct = 0.25, logfc.threshold = 0.5)

nrow(markers)
head(markers)
```

```{r top-markers}
top5 <- markers |>
    group_by(cluster) |>
    slice_max(avg_log2FC, n = 5) |>
    summarise(top_genes = paste(gene, collapse = ", "))
top5
```

## 1.1 Canonical markers

Published markers for the major cardiac populations (Sim et al. 2021 and
standard references). Dot size is the fraction of nuclei expressing the gene;
colour is mean expression.

```{r canonical-panel}
panel <- list(
    Cardiomyocyte  = c("TNNT2", "ACTN2", "TNNI3", "RYR2", "MYH6", "TTN"),
    Fibroblast     = c("COL1A1", "COL3A1", "DCN", "PDGFRA"),
    Endothelial    = c("PECAM1", "CDH5", "VWF", "FLT1"),
    Immune         = c("PTPRC", "CD68", "CD163", "CD74"),
    Epicardial     = c("UPK3B", "WT1", "MSLN"),
    Neural         = c("PLP1", "S100B", "NRXN1"),
    `Smooth muscle`= c("MYH11", "TAGLN", "MYLK")
)
genes <- unlist(panel, use.names = FALSE)
genes <- genes[genes %in% rownames(seu)]   # drop anything filtered out earlier
setdiff(unlist(panel, use.names = FALSE), genes)  # what was missing, if any
```

```{r dotplot, fig.width=13, fig.height=6}
DotPlot(seu, features = genes, assay = "SCT") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(x = NULL, y = "cluster")
```

---

# 2. Naming the clusters

> ### <span style="color:#C0392B">DECISION 4 — what is your cluster? *(15 min)*</span>
>
> Your demonstrator will give each pair **one cluster number**.
>
> 1. Pull its top markers with the chunk below.
> 2. Look them up. [PanglaoDB](https://panglaodb.se/search.html) is a searchable
>    database of cell-type markers; the [Human Protein Atlas](https://www.proteinatlas.org/)
>    is good for single genes.
> 3. **Then go back and check your cluster in the DotPlot above.** 
>    `FindAllMarkers` returns what makes your cluster *different from the
>    other eighteen* — which is not the same as what makes it what it is. If
>    eleven clusters are the same cell type, the genes defining that cell type
>    cancel out and never appear in any of their marker lists.
> 4. **Commit to an answer before you look at the reveal.** Write down the cell
>    type *and* your confidence: certain / fairly sure / guessing.

```{r my-cluster}
my_cluster <- 5   # <- YOUR CLUSTER NUMBER

markers |>
    filter(cluster == my_cluster) |>
    slice_max(avg_log2FC, n = 15) |>
    select(gene, avg_log2FC, pct.1, pct.2, p_val_adj)
```

```{r my-cluster-umap, fig.width=11, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
p2 <- DimPlot(seu, reduction = "umap",
              cells.highlight = colnames(seu)[seu$seurat_clusters == my_cluster],
              cols.highlight = "#C0392B") +
    NoLegend() + ggtitle(paste("Cluster", my_cluster))
p1 + p2
```

## 2.1 The reveal

These are the labels from the published Sim et al. analysis.

> ### <span style="color:#C0392B">"Published" is not "correct"</span>
>
> These labels are one careful team's interpretation, produced by people who
> faced the same choices you just did. In at least two clusters here a
> marker-based reading is *more specific* than the published label, and in at
> least one the markers point somewhere the label does not.
>
> **If your answer disagrees, work out why before assuming you were wrong.**

```{r reveal}
key <- readRDS(file.path(DATA, "celltype_key.rds"))
tab <- table(cluster = seu$seurat_clusters, published = key[colnames(seu)])

# For your cluster: what did the original authors call these cells?
sort(tab[as.character(my_cluster), ], decreasing = TRUE) |> head(3)
```

```{r reveal-all}
data.frame(
    cluster   = rownames(tab),
    n         = as.integer(rowSums(tab)),
    published = colnames(tab)[apply(tab, 1, which.max)],
    agreement = paste0(round(100 * apply(tab, 1, max) / rowSums(tab)), "%")
)
```

**Look at that table as a whole before moving on.** We produced **19 clusters**, and **eleven of them are cardiomyocytes**. That is not a failure — cardiomyocytes genuinely differ across
foetal, young and adult hearts, and the algorithm is finding real structure. 

**Too coarse.** The published analysis describes **8** cell types. Your 19
clusters recover **6**. Run this:

```{r rare-types}
key_tab <- sort(table(key[colnames(seu)]), decreasing = TRUE)
key_tab
```

**Smooth muscle cells** (62 nuclei) and **erythroid cells** (9 nuclei) are in
your data, and no cluster corresponds to either. Sixty of the 62 smooth muscle
nuclei are sitting inside cluster 3 — which is why cluster 3's agreement with the
published labels is the worst:

```{r smooth-muscle}
sm <- names(key)[key == "Smooth muscle cells"]
table(seu$seurat_clusters[colnames(seu) %in% sm])
```

So the same analysis simultaneously split one cell type into eleven and merged
two others into invisibility. **Nothing in the output flagged either.** Rare
populations are the ones most likely to be lost, and — as you saw with the QC
thresholds in Practical 1 — also the ones most likely to be filtered away
earlier. Both failures point the same direction.

> ### <span style="color:#2980B9">🤖 What the agent found here</span>
>
> Claude Science scored **eleven** canonical categories against its clusters:
> cardiomyocyte, endothelial, fibroblast, pericyte/SMC, myeloid, lymphoid, mast,
> neural/glial, epicardial, erythroid, cycling.
>
> Note what is on that list that is **not** in your six: pericyte/smooth muscle,
> mast cells, erythroid — and it split immune into myeloid vs lymphoid, which is
> your clusters 6 and 16.
>
> It looked for the rare populations *by name*, because it started from a
> reference panel rather than from the clusters. You started from the clusters
> and asked what they were. 

> **<span style="color:#C0392B">HARD SYNC POINT.</span>** Run this now regardless of
> where you got to. It applies the reference labels so everyone's downstream
> numbers match.

```{r sync-2}
labels <- readRDS(file.path(DATA, "cluster_labels.rds"))
seu$celltype <- unname(labels[as.character(seu$seurat_clusters)])
table(seu$celltype)
```

```{r annotated-umap, fig.width=12, fig.height=5}
p1 <- DimPlot(seu, reduction = "umap", group.by = "celltype", label = TRUE,
              repel = TRUE) + NoLegend() + ggtitle("Cell type")
p2 <- DimPlot(seu, reduction = "umap", group.by = "group", cols = group_cols,
              shuffle = TRUE) + ggtitle("Developmental stage")
p1 + p2
```

---

# 3. Does composition change with development?

The first biological question: as the heart matures, do the *proportions* of cell
types change?

```{r composition, fig.width=10, fig.height=5}
comp <- seu@meta.data |>
    count(sample, group, celltype) |>
    group_by(sample) |>
    mutate(prop = n / sum(n)) |>
    ungroup()

ggplot(comp, aes(sample, prop, fill = celltype)) +
    geom_col() +
    facet_grid(~ group, scales = "free_x", space = "free_x") +
    labs(x = NULL, y = "proportion of nuclei", fill = NULL) +
    theme_minimal()
```

```{r composition-box, fig.width=11, fig.height=5}
ggplot(comp, aes(group, prop, fill = group)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, size = 1.2) +
    facet_wrap(~ celltype, scales = "free_y", nrow = 2) +
    scale_fill_manual(values = group_cols) +
    labs(x = NULL, y = "proportion") +
    theme_minimal() + theme(legend.position = "none")
```


We use `propeller` to test if proportions change across developmental stages: it computes each donor's proportions, transforms
them (proportions are bounded and heteroscedastic), and tests across donors.

```{r propeller}
prop_res <- propeller(clusters = seu$celltype, sample = seu$sample,
                      group = seu$group)
prop_res
```

This is a strong result even at n = 3 per group. Cardiomyocytes fall from about
**74% of nuclei in foetal hearts to 24% in adult**; immune cells and fibroblasts
rise correspondingly. Neurons do not change.

One caution: **Compositions are constrained to sum to 1.** If cardiomyocytes fall, then *everything else must rise* whether or not anything happened to it. You cannot read the fibroblast result independently of the cardiomyocyte result.

Recall too from Practical 1 that donor `a3` contributes far fewer nuclei than the others, so its proportions are the noisiest of the nine.

> ### <span style="color:#2980B9">🤖 What the agent did here</span>
>
> It reported cardiomyocytes falling from **64.4%** of cells in foetal hearts to
> **23.3%** in adult — essentially the same numbers you just produced — and
> described the change as developmental biology.
>
> **It ran no statistical test of composition at all.** 

---

# 4. Differential expression 

Second question: within cardiomyocytes, which genes change between foetal and adult hearts?

## 4.1 The naive approach

Every scRNA-seq tutorial reaches for `FindMarkers()`. Let us run it — but on a **random 100 nuclei per group**, so it finishes in about ten seconds.

```{r cm-subset}
cm <- subset(seu, subset = celltype == "Cardiomyocytes" &
                           group %in% c("adult", "fetal"))
table(cm$group, cm$sample)
```

```{r naive-de-small} 
set.seed(42) # This is AI's favourite random seed, guess why
cells <- unlist(lapply(c("adult", "fetal"), function(g)
    sample(colnames(cm)[cm$group == g], 100)))
small <- subset(cm, cells = cells)

Idents(small) <- "group"
naive_small <- FindMarkers(small, ident.1 = "adult", ident.2 = "fetal",
                           assay = "SCT", verbose = FALSE)
sum(naive_small$p_val_adj < 0.05)
```

Several hundred significant genes, from 200 nuclei.

Now: If we repeat this with all
~3,200 cardiomyocytes instead of 200, what happens to that number? 

## 4.2 Why so many false positives 

Take two foetal donors. Same developmental stage, same tissue, same protocol. There is no developmental difference between `f1` and `f2`.

These are pre-computed (about four minutes to run), all at the **same 300 cells per side**, so the numbers are directly comparable.

```{r de-demo}
demo <- readRDS(file.path(DATA, "de_demo.rds"))
demo$wilcoxon
```

```{r demo-plot, fig.width=7.5, fig.height=4}
ggplot(demo$wilcoxon, aes(reorder(comparison, n_sig), n_sig, fill = kind)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(`real contrast` = "#2980B9",
                                 `same stage (foetal)` = "#C0392B")) +
    labs(x = NULL, y = "genes called significant (padj < 0.05)", fill = NULL,
         title = "Per-nucleus Wilcoxon, 300 cells per side",
         subtitle = "the red bars are comparisons with no biology in them") +
    theme_minimal()
```


## How should we interpret the results?



## 4.3 More defensible approach: pseudobulk

It is generally acceptable to assume that different donors are independent, even within the same condition. Hence we create **pseudo-bulks** from single nuclei.

`Seurat2PB()` **sums the raw counts** across all nuclei of one cell type within one donor. One column per donor × cell type, on the same count scale as a bulk RNA-seq library — so the whole bulk toolkit (edgeR/limma) applies unchanged, and the replicate is now the donor rather than the nucleus. 

```{r pseudobulk}
seu$Celltype <- seu$celltype
seu$Sample   <- seu$sample
DefaultAssay(seu) <- "RNA"

pb <- Seurat2PB(seu, sample = "Sample", cluster = "Celltype")
dim(pb)

# Cardiomyocytes only
pb_cm <- pb[, pb$samples$cluster == "Cardiomyocytes"]
pb_cm$samples$group <- sub("[0-9]+$", "", pb_cm$samples$sample)
pb_cm$samples$group <- c(a = "adult", f = "fetal", y = "young")[pb_cm$samples$group]
table(pb_cm$samples$group)
```

Nine columns. That is the real sample size, and it was always the real sample
size.

The rest is a classic bulk limma-voom pipeline:

- **`filterByExpr()`** drops genes with too few counts to be testable at n = 9 — untestable genes only cost multiple-testing power.
- **`normLibSizes()`** computes TMM scaling factors, so that a donor whose library is dominated by a few very high genes is not mistaken for one where everything else is lower.
- **`voom()`** converts counts to log-CPM and gives each observation a **precision weight**. Count data are heteroscedastic: low-expressed genes are noisier than high-expressed ones. `voom` estimates that mean–variance trend (the plot) and down-weights the noisy observations, so ordinary linear models can then be used on the log scale.
- **`lmFit()` + `contrasts.fit()`** fit a linear model per gene and extract the comparison we care about, `adult - fetal`.
- **`eBayes()`** — *empirical Bayes*. With only 3–4 donors per group, a gene's own variance estimate is unreliable, and a gene that is noisy by chance produces a tiny p-value. `eBayes` shrinks each gene's variance towards the trend shared by all genes, giving a **moderated t-statistic**: borrowing information across genes buys back degrees of freedom we do not have in the donors.

```{r pb-de}
keep <- filterByExpr(pb_cm, group = pb_cm$samples$group)
pb_cm <- pb_cm[keep, , keep.lib.sizes = FALSE]
pb_cm <- normLibSizes(pb_cm)

design <- model.matrix(~ 0 + factor(pb_cm$samples$group))
colnames(design) <- levels(factor(pb_cm$samples$group))

v   <- voom(pb_cm, design, plot = TRUE)
fit <- lmFit(v, design)
fit <- contrasts.fit(fit, makeContrasts(adult - fetal, levels = design))
fit <- eBayes(fit)

summary(decideTests(fit))
```

```{r pb-top}
topTable(fit, n = 15)
```

```{r volcano, fig.width=7, fig.height=5.5}
tt <- topTable(fit, n = Inf)
tt$sig <- tt$adj.P.Val < 0.05 & abs(tt$logFC) > 1
ggplot(tt, aes(logFC, -log10(adj.P.Val), colour = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#C0392B")) +
    labs(title = "Cardiomyocytes: adult vs foetal (pseudobulk)",
         x = "log2 fold change", y = "-log10 adjusted p") +
    theme_minimal() + theme(legend.position = "none")
```

> ### <span style="color:#2980B9">🤖 What the agent did here</span>
>
> Same contrast, same data. It ran **per-cell Wilcoxon** and reported
> **5,967 significant genes** as its headline result, with a volcano plot.
>
> Then, in its caveats:
>
> > *"Wilcoxon on individual cells treats cells as replicates and **inflates
> > significance** — for anything you'd report, redo this as donor-level
> > pseudobulk with DESeq2 at n=3 per stage."*
>
>
> **But it did not do it.** Pseudobulk appears in its output only as a heatmap. 




```{r session}
sessionInfo()
```
