---
title: "BINF90004 Single cell RNA-seq case study"
author: "Dr Jiadong Mao"
date: "Sep 2023"
output: html_document
fontsize: 10pt
---

## Welcome!

<!-- ## \red{Notation} -->

<!-- - $\p$: probability -->

<!-- - $p$: pmf pdf, never prob -->

<!-- - $\theta$: parameter, may be vector, eg $\theta=(\mu,\sigma^2)$ -->

<!-- - $\sample = (Y_1,\dots,Y_n)$, sample of size $n$ viewed as rvs -->

<!-- - $Y$: a generic $Y_i$ (iid as $\sample$) -->

<!-- - $\bmy = \{y_1,\dots,y_n\}$, sample of size $n$, observed (nonrandom) -->

<!-- - $y$: a generic $y_i$ -->

-   About me.

-   About you.

## Course information

-   Lectures

    -   1.5h on Mon & Wed; 1h on Tue & Thur;
    -   Lab session will be incorporated into the lectures.

-   Consultation hours

    -   Your preference?

-   Q&A out of consultation hours: Ed.

-   3 quizzes, 5 assignments: goal is to monitor learning, will involve R programming (no more than what appeared in the lectures).

-   Mid-term exam, final exam: mainly about understanding, some derivagions, no programming.

-   Any question?

## Course overview

```{r, echo=T}
knitr::opts_chunk$set(cache=T, message = F, echo=T)
```

-   What the course is about?
    -   Bayesian inference: a Bayesian version of undergraduate level mathematical statistics;
    -   Topics include: fitting distributions to data, group comparisons, linear models, generalised linear models;
    -   Learn to appreciate advantages of Bayesian approaches.
-   Mathematics: will involve a lot of derivations; but much has been seen in prerequisite courses.

## Example (contd)

```{r, echo=F, message=F, fig.dim=c(10,7), cache=T}
library(tidyverse)
plot_beta <- tibble(theta=seq(0,1,length.out=200),
                    'beta(1,1)'=dbeta(theta,1,1),
                    'beta(0.5,0.5)'=dbeta(theta,0.5,0.5),
                    'beta(2,2)'=dbeta(theta,2,2),
                    'beta(1,5)'=dbeta(theta,1,5),
                    'beta(5,1)'=dbeta(theta,5,1),
                    'beta(2,5)'=dbeta(theta,2,5))
plot_beta %>% 
  pivot_longer(cols=starts_with('beta')) %>% 
  ggplot(aes(x=theta,y=value)) + 
  geom_line(aes(colour=name), size=1)
```

## Posterior

-   How data change belief: eg $a=2, b=20, y=0$

```{r, echo=F, fig.dim=c(8,5), cache=T}
plot_beta <- tibble(theta=seq(0,1,length.out=200),
                    beta1=dbeta(theta,2,20),
                    beta2=dbeta(theta,2,40))
plot_beta %>% 
  pivot_longer(cols=starts_with('beta')) %>% 
  ggplot(aes(x=theta,y=value)) + 
  geom_line(aes(colour=name))
```
